---
title: 'Thermal performance models for *Lemna minor*'
subtitle: 'Response variable: RGR'
author: "Debora"
date: "4/14/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Installing packages and enabling libraries
```{r message=FALSE, warning=FALSE}
library(rTPC)
library(ggplot2)
library(nls.multstart)
library(broom)
library(dplyr)
library(AICcmodavg)
library(tidyverse)
```

### Download data and subset to include only performance and temperature
```{r}
duckweed_constant <- read.csv("https://raw.githubusercontent.com//Cuddington-Lab/thermal-experiments/main/duckweed_constant.csv",
                     header=TRUE)
duckweed_constant <- subset(duckweed_constant, select = c(Mean_Temp, RGR))
```

### Fit the chosen models to the duckweed data
```{r}
label_facets_num <- function(string){
  len <- length(string)
  string = paste('(', 1:len, ') ', string, sep = '')
  return(string)
}

d_fits <- nest(duckweed_constant, data = c(Mean_Temp, RGR)) %>%
  mutate(
    ratkowsky = map(data, ~nls_multstart(RGR~ratkowsky_1983(temp = Mean_Temp, tmin, 
    tmax, a, b), data = duckweed_constant, iter = c(4,4,4,4),
    start_lower = get_start_vals(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'ratkowsky_1983') - 1,
    start_upper = get_start_vals(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'ratkowsky_1983') + 1,
    lower = get_lower_lims(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'ratkowsky_1983'), upper = get_upper_lims(duckweed_constant$Mean_Temp, 
    duckweed_constant$RGR, model_name = 'ratkowsky_1983'), supp_errors = 'Y',
    convergence_count = FALSE)),
    
    briere2 = map(data, ~nls_multstart(RGR~briere2_1999(temp = Mean_Temp,tmin, tmax, a, b),
    data = duckweed_constant, iter = c(4,4,4,4), 
    start_lower = get_start_vals(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'briere2_1999') - 1, 
    start_upper = get_start_vals(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'briere2_1999') + 1, lower = get_lower_lims(duckweed_constant$Mean_Temp, 
    duckweed_constant$RGR, model_name = 'briere2_1999'), 
    upper = get_upper_lims(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'briere2_1999'), supp_errors = 'Y', convergence_count = FALSE)),
    
    gaussian = map(data, ~nls_multstart(RGR~gaussian_1987(temp = Mean_Temp, rmax, topt, a),
    data = duckweed_constant, iter = c(4,4,4),
    start_lower = get_start_vals(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'gaussian_1987') - 1,
    start_upper = get_start_vals(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'gaussian_1987') + 1,
    lower = get_lower_lims(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'gaussian_1987'), upper = get_upper_lims(duckweed_constant$Mean_Temp, 
    duckweed_constant$RGR, model_name = 'gaussian_1987'),
    supp_errors = 'Y',
    convergence_count = FALSE)),
    
    lactin2 = map(data, ~nls_multstart(RGR~lactin2_1995(temp = Mean_Temp, a, b, tmax, delta_t),
    data = duckweed_constant, iter = c(4,4,4,4),
    start_lower = get_start_vals(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'lactin2_1995')-1,
    start_upper = get_start_vals(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'lactin2_1995')+1, lower = get_lower_lims(duckweed_constant$Mean_Temp, 
    duckweed_constant$RGR, model_name = 'lactin2_1995'),
    upper = get_upper_lims(duckweed_constant$Mean_Temp, duckweed_constant$RGR, 
    model_name = 'lactin2_1995'), supp_errors = 'Y', convergence_count = FALSE))
    )
```

### Stack models
```{r}
d_stack <- select(d_fits, -data) %>%
  pivot_longer(., names_to = 'model_name', values_to = 'fit', ratkowsky:lactin2)
```

### Obtain parameters
```{r}
params <- d_stack %>%
  mutate(., est = map(fit, tidy)) %>%
  select(-fit) %>%
  unnest(est)
```

### Obtain model predictions
```{r}
newdata <- tibble(Mean_Temp = seq(min(duckweed_constant$Mean_Temp), max(duckweed_constant$Mean_Temp), length.out = 60))

d_preds <- d_stack %>%
  mutate(., preds = map(fit, augment, newdata = newdata)) %>%
  select(-fit) %>%
  unnest(preds)

params1 <- d_stack %>%
  mutate(., params = map(fit, calc_params)) %>%
  select(-fit) %>%
  unnest(params)
estimates <- bind_rows(select(params1, model_name, rmax:skewness), params1) %>%
  mutate_if(is.numeric, round, 2)
estimates[!duplicated(estimates$model_name), ]
```
   
### Plot model fits
```{r}
ggplot(d_preds) +
  geom_line(aes(Mean_Temp, .fitted, col = model_name),size = 1.5) +
  geom_point(aes(Mean_Temp, RGR), duckweed_constant) +
    labs(x = "Mean_Temperature (ºC)",
       y = "Relative growth rate (RGR)",
       title = "Fitted thermal performance curves")
```                     

### Model assessment
```{r}
d_ic <- d_stack %>%
  mutate(., info = map(fit, glance),
         AICc =  map_dbl(fit, MuMIn::AICc)) %>%
  select(-fit) %>%
  unnest(info) %>%
  select(model_name, AICc)
d_ic <- distinct(d_ic)
d_ic %>% arrange(AICc)
```

### Reference
# Padfield, D., O’Sullivan, H., & Pawar, S. (2021). rTPC and nls.multstart: A new pipeline to fit thermal 
# performance curves in R. Methods in Ecology and Evolution. https://doi.org/10.1111/2041-210X.13585