---
title: 'Anova for the pea-aphid experiment at 15 C - autocorrelated temperatures '
author: "Jon, Champ, Sunandan, Debora"
date: "10/08/2021"
output: pdf_document
---
### Purpose
To determine if there is a significant relationship between aphid offspring (dependent variable) and different autocorrelated temperature regimes (independent variable)


### Installing packages and enabling libraries
```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
```

### Load metafile data from GitHub (expdata_metafile_01062021.csv)
### and attribute columns as factors or numeric, as appropriate
```{r message=FALSE, warning=FALSE}
aphid.offspring <- read.csv("C:/Users/user/Desktop/uWaterloo/experiments/expdata_metafile_01062021.csv",
header = TRUE, colClasses = c("factor", "factor","factor","factor","factor",
                                "factor","factor","numeric","numeric","numeric",
                                "numeric","numeric","numeric","factor"))
```

### Create new data frame including only mean temp of 15C
```{r}
anova.15C <- filter(aphid.offspring, Mean_Temp == 15)
```

### Remove 0.9 rows (Julie's recommendation)
```{r}
anova.15C <- anova.15C[-c(2, 4, 7, 8), ] 
```

### Concatenate columns Autocorrelation and Cat to enable autocorrelation regime
### labelling (new column name: Autocorr_label)
```{r}
anova.15C$Autocorr_label<-paste(anova.15C$Autocorrelation,anova.15C$Cat)
```

### Rename autocorrelation regimes 
```{r}
anova.15C$Autocorr_label[anova.15C$Autocorr_label == "0.95 N"] <- "a=0.95,m<0"
anova.15C$Autocorr_label[anova.15C$Autocorr_label == "0.95 N/A"] <- "a=0.95,m<0"
anova.15C$Autocorr_label[anova.15C$Autocorr_label == "0.9 N"] <- "a=0.95,m<0"
anova.15C$Autocorr_label[anova.15C$Autocorr_label == "0.9 N/A"] <- "a=0.95,m<0"
anova.15C$Autocorr_label[anova.15C$Autocorr_label == "0.6 N/A"] <- "a=0.6"
anova.15C$Autocorr_label[anova.15C$Autocorr_label == "0 N/A"] <- "a=0"
anova.15C$Autocorr_label[anova.15C$Autocorr_label == "0.95 P"] <- "a=0.95,m>0"
```

### Create new column including sum of offspring (Total_Offspring) 
```{r}
anova.15C$Total_Offspring <- anova.15C$Offspring_Plant1 + 
  anova.15C$Offspring_Plant2 + anova.15C$Offspring_Plant3
```

### Run an analysis of variance
```{r}
res.aov <- aov(Total_Offspring ~ Autocorr_label, data = anova.15C)
summary(res.aov)
```

### Boxplots
```{r}
min.mean.sd.max <- function(x) {
  r <- c(min(x), mean(x) - sd(x), mean(x), mean(x) + sd(x), max(x))
  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
  r
}
boxplots <- ggplot(aes(y = Total_Offspring, x = factor(Autocorr_label)), 
                   data = anova.15C)
boxplots <- boxplots + stat_summary(fun.data = min.mean.sd.max, 
                                    geom = "boxplot") + 
  geom_jitter(position=position_jitter(width=.2), size=3) + 
  ggtitle("Mean temperature 15°C") + 
  xlab("Autocorrelation") + ylab("Total_Offspring")
boxplots
```
   
### Linear model (0 autocorrelation vs. other groups)
```{r}
lm.15C <- lm(Total_Offspring ~ Autocorr_label, data = anova.15C)
summary(lm.15C)
```                     
