
---
title: "Survival effects of strong autocorrelation and hot average temperatures"
author: "Debora"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## H2: Effect of initial conditions
  
When temperatures are strongly autocorrelated and average is higher than optimal:  
  
- Hot start: negative impact to performance  
- Cold start: positive impact to performance  
  
This is not observed in other average temperatures (due to the asymmetrical shape of thermal performance curves: prolonged exposures to temperatures either above or below the average typical of strongly autocorrelated thermal regimes are not as detrimental as average temperatures approaching the upper thermal limit) 

### Statistical analyses
Survival: binary categorization of fronds (Binomial distribution)  
1 = living frond  
2 = dead frond  

- Generalized linear mixed model (logistic regression) 
- Fixed factors: autocorrelation strength (coefficient = 0 or 0.95 starting hot or 0.95 starting cold)  
- Random factors: experiment number (nesting of experiment runs; passage of time) and incubator   
- Model selection (Akaike weights) and model diagnostics for selected model  
- Post-hoc tests: least-squares means with Satterthwaite approximations for df  

```{r include=FALSE}
#Overview of dataset: data we typically collect in a given experimental run (5 days)

### This dataset contains replicates for which a preparation technique was performed (repeated first born frond selection to reduce maternal effects)
original_dataset_2 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2022-2023.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")
original_dataset_2$prep <- rep("yes",times=length(original_dataset_2$Experiment_Number))

### This dataset contains replicates of experiments performed without a preparation technique 
original_dataset_1 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2021-2022.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")
original_dataset_1$prep <- rep("no",times=length(original_dataset_1$Experiment_Number))

datin <- rbind(original_dataset_1,original_dataset_2)
```

```{r include=FALSE}
## Data validation and data cleaning
#Excluding failed experiments outside established parameters:  
  
#* Experiment duration (120h +- 1h)  
#* Standard deviation (2.5ºC +- 0.4ºC)  
#* Autocorrelation (ρ=0 or ρ=0.95 +-0.3)  
### Data validation: excluding experiments outside thresholds established in my methods section (for standard deviation and autocorrelation of observed temperatures)
datin <- datin[!(datin$Treatment == 0 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9))
               &!(datin$Treatment == 0.95 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9)),]

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_ac <= -0.2 | datin$Obs_ac >= 0.2))
               &!(datin$Treatment == 0.95 & (datin$Obs_ac <= 0.92 | datin$Obs_ac >= 0.98)),]
```

```{r echo=FALSE}
### Excluding failed experiments
datin <- subset(datin, !Errors == "y"|is.na(Errors))
```

```{r echo=FALSE}
### Excluding invasive species (I am focusing on L. minor for this analysis)
datin <- subset(datin,!Species=="LP")
### Excluding constant temperature experiments
datin <- subset(datin, !Treatment == "constant")
```

```{r include=FALSE}
datin$Frond_count <- datin$Frond_count_1 + datin$Frond_count_2 + datin$Frond_count_3
```

```{r include=FALSE}
datin <- datin[!is.na(datin$Experiment_Number),]

### Create new treatment
table(datin$Treatment, datin$cat_1)
levels(datin$cat_1) = c("","N","N/A","P")
datin$label<-paste0(datin$Treatment, datin$cat_1)
table(datin$label)

colnames(datin)[6] <- "Treatment1"
colnames(datin)[34] <- "Treatment"

### Relabel groups to simplify visualization of results
library(stringr)
datin$Treatment <- str_replace(datin$Treatment, "0N/A", "0")
datin$Treatment <- str_replace(datin$Treatment, "0.95N", "hot-cold")
datin$Treatment <- str_replace(datin$Treatment, "0.95P", "cold-hot")
datin$Treatment <- str_replace(datin$Treatment, "0", "no autocorrelation")
datin$Mean_temperature <- str_replace(datin$Mean_temperature, "15", "10")
datin$Mean_temperature <- str_replace(datin$Mean_temperature, "10", "10-15")
```

### Dataset: each row of the dataset corresponds to a single frond which either survived (1) or did not survive (0) at the end of the experiment
```{r echo=FALSE}
## H2: Survival at 37 ºC
#hot-cold x no autocorrelation and cold-hot x no autocorrelation: p<0.001; hot-cold x cold-hot: p=0.03  
#Marginal R² / Conditional R²:	0.373 / 0.711 
datins <- subset(datin,Mean_temperature=="37")

datins$Mortality <- datins$Dead_frond_count_1 + datins$Dead_frond_count_2 + datins$Dead_frond_count_3

mort <- as.data.frame(lapply(datins, rep, datins$Mortality))
mort <- subset(mort, !mort$Mortality == 0)
mort["Mortality"][mort["Mortality"] >= 1] <- 0
names(mort)[names(mort) == 'Mortality'] <- 'survival'
mort <- subset(mort, select = -(Frond_count))

living <- as.data.frame(lapply(datins, rep, datins$Frond_count))
living <- subset(living, !living$Frond_count == 0)
living["Frond_count"][living["Frond_count"] >= 1] <- 1
names(living)[names(living) == 'Frond_count'] <- 'survival'
living <- subset(living, select = -(Mortality))

datins2 <- rbind(living,mort)

knitr::kable(datins2[c(120:123,975:979), c(1,5,33,34)],row.names = FALSE)
```

### Plotting survival data: each dot also corresponds to a single individual
1 = living frond  
0 = dead frond  
  
```{r echo=FALSE}
library(ggplot2)
ggplot(datins2, aes(x = Treatment, y = survival, color = Treatment)) + 
  theme_bw() +
  geom_jitter() +
  xlab("Treatment") +
  ggtitle("Duckweed survival at 37 ºC") +
  ylab("Survival") +
  geom_smooth(formula = y ~ x,method=lm,se=T) + theme_bw(base_size=34) +
  theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(), 
          axis.line = element_line(colour = "black")) +
  scale_color_manual(values = c(" no autocorrelation" = "grey37",
                                "hot-cold" = "#D55E00", "cold-hot" = "#0072B2")) +
  scale_y_continuous(breaks = seq(0, 1, len = 2)) +
  theme(legend.position = "none",axis.text=element_text(size=14),
        axis.title=element_text(size=14,face="bold"), plot.title = element_text(size=20))
```

### Candidate models (logistic fit, binomial data)
```{r echo=TRUE, message=FALSE, warning=FALSE, comment=NA}
library(sjPlot)
library(sjmisc)

datins2$Experiment_Number <- as.factor(datins2$Experiment_Number)
datins2$Incubator <- as.factor(datins2$Incubator)

datins2 <- within(datins2, Treatment <- relevel(factor(Treatment), ref = "no autocorrelation"))

# Logistic regression allows categorically and continuously scaled variables to predict any categorically 
# scaled criterion (Osborne, J. W. (Ed.). 2008. Best practices in quantitative methods. Sage)

# adding random effects
library(lme4)
glm.fit.sp <- glm(survival ~ Treatment, family = binomial, data=datins2)
experim <- glmer(survival ~ Treatment + (1|Experiment_Number), family=binomial, data=datins2)
incub <- glmer(survival ~ Treatment + (1|Incubator), family=binomial, data=datins2)
inc_exp <- glmer(survival ~ Treatment + (1|Incubator)+(1|Experiment_Number), family=binomial, data=datins2)
exp.inc <- glmer(survival ~ Treatment + (1|Experiment_Number/Incubator), family=binomial, data=datins2)
```


### Model selection: The best model includes incubator number nested in experiment number as a random effect 
```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NA}
Cand.mods <- list("no random effects" = glm.fit.sp,
                  "experiment number" = experim,
                  "incubator" = incub, 
                  "incubator and experiment number" = inc_exp,
                  "incubator nested in experiment number" = exp.inc)

library(performance)
compare <- compare_performance(Cand.mods)
knitr::kable(compare[,c(1,2,3,4)],"simple", digits = 2)
```  

### Statistical analysis of selected model  
- Model assumptions for over dispersion and normality of random effects were met
- Odds ratio: probability of success / probability of failure
- For categorical predictors, the odds ratio compares the odds of the event occurring for each category of the predictor relative to the reference category, given that all other variables remain constant

```{r echo=FALSE, message=FALSE, warning=FALSE}
tab_model(exp.inc)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
### model assumptions check
#to check for over dispersion: the variance of the response is greater than what's assumed by the model
library(spida2)
overdisp_fun(exp.inc)
#Ratio is not significantly > 1: variance is not increasing more rapidly than the mean

## for glm = random effects are normally distributed!!
r_int <- ranef(exp.inc)$Experiment_Number$`(Intercept)`
qqnorm(r_int)
qqline(r_int)
shapiro.test(r_int)
```

### Area under the curve 
- From 0 to 1: refers to the ability of the model to correctly distinguish between the 2 classes
```{r echo=TRUE, message=FALSE, warning=FALSE}
#AUC
glm.probs <- predict(exp.inc,datins2, type = "response")
### Making a receiver operator curve (ROC) to analyze the model performance across several thresholds 
library(ROCit)
ROCit_obj <- rocit(score = glm.fit.sp$fitted.values, class = datins2$survival)
plot(ROCit_obj)

### Obtaining AUC - area under the curve (maximum is one) 
summary(ROCit_obj) 
```

Sensitivity: proportion of true positives (correctly classified)
Specificity: proportion of true negatives (correctly classified)
```{r echo=TRUE, message=FALSE, warning=FALSE}
### Optimal threshold value given by the Youden index 
library(pROC)
roc_obj <- roc(datins2$survival, glm.probs)
coords(roc_obj, "best", "threshold")
```

### Plotting model predictions
```{r echo=FALSE, message=FALSE, warning=FALSE}
my_title <- expression(paste("Predicted ",italic("L. minor")," survival at 37 °C"))
library(ggeffects)
library(dplyr)
library(ggplot2)
ggpredict(exp.inc, "Treatment") %>% plot(dot.size = 5,line.size = 2.5) + 
  theme_bw(base_size=20) +
  ggtitle ("Predicted survival at 37 °C") +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = "Survival probability", x = "Treatment") +
  ggplot2::scale_color_manual(values=c('#999999','#E69F00',"black"), aesthetics = "colour")
```

### Post-hoc analyses
```{r message=FALSE, include=FALSE}
library(emmeans) #emmeans has been used for GLMER & binary data (internet forums)
pairwise <- emmeans(exp.inc, list(pairwise ~ Treatment), lmer.df="satterthwaite")
pairwise
pairs <- as.data.frame(pairwise$`pairwise differences of Treatment`)
#pairs <- subset(pairs,p.value<=0.05)
options(scipen = 999)
knitr::kable(pairs,"simple",digits=4)
```

```{r echo=FALSE, message=FALSE}
knitr::kable(pairs,"simple",digits=4)
```
