---
title: "Dotted duckweed maxent study: Preliminary model run (rmd file #1)"
author: "Debora"
date: "`r Sys.Date()`"
output: html_document
---

<style type="text/css">
  body{
  font-size: 15pt;
}
</style>

### Purpose: To create a preliminary model using pre-selected bioclimatic variables. I will check for correlations and importance of each variable to the model to perform a subsequent variable selection which will include the most important variables with the lowest levels of correlation between them.

```{r message=FALSE, warning=FALSE, include=FALSE}
Sys.setenv('R_MAX_VSIZE'=32000000000)
memory.limit(240000)

options(java.parameters="~xmx10g")
library(raster)
library(maptools)
library(dismo)
library(maps)
library(rJava)
.jinit()
```

### Obtain occurrence records from GBIF and remove duplicates and records missing information
Downloaded from GBIF on Nov 29, 2023
https://doi.org/10.15468/dl.x7svjw
```{r echo=TRUE, message=FALSE, warning=FALSE}
setwd("C:/Users/user/Desktop/maxent_polished")

lp <- read.csv("lp.csv")
head(lp[,c(14,22,23)])
```

### Clean up dataset 
(removing data without coordinates and duplicates)
```{r echo=TRUE, message=FALSE, warning=FALSE}
#removing data without geographic coordinates
lp <- subset(lp, !is.na(lon) & !is.na(lat))

#removing duplicates
dups <- duplicated(lp[, 1:10])
lp_clean  <-  lp[dups, ]
```

### Add occurrences from literature 
(for cases where coordinates are not exact, coordinates of a given country's capital were used)
```{r echo=TRUE, message=FALSE, warning=FALSE}
#Exact records obtained from published papers
europe_lit <- data.frame(lon=c(4.51,4.95,4.31,4.25,6.12,5.23,16.15,12.27,12.45),lat=c(52.18,52.23,52,52,51.31,51.48,28.28,43.56,41.9))

#Inexact records obtained from published papers
europe_lit_unreported <- data.frame(lon=c(4.47,0.11,12.57,12.57,6.14,24.94,18.3),lat=c(50.5,51.5,41.87,41.87,46.2,60.19,59.2))
```

### Create datasets for each geographic location
```{r echo=TRUE, message=FALSE, warning=FALSE}
#subsetting by geographical region
world <- subset(lp_clean, continent != "EUROPE")
world <- world[, c("lon", "lat")]

europe_GBIF <- subset(lp_clean, continent == "EUROPE")
europe_GBIF <- europe_GBIF[, c("lon", "lat")]

#European dataset containing all records
europe <- rbind(europe_GBIF,europe_lit,europe_lit_unreported)

#European dataset including only exact records
europe_sure <- rbind(europe_GBIF,europe_lit)
```

### Plot cleaned dataset of occurrences (GBIF and literature records)
```{r echo=FALSE, message=FALSE, warning=FALSE}
allrecords <- rbind(europe,world)
data(wrld_simpl)
plot(wrld_simpl, col = "light gray", axes = T, main = "L. punctata GBIF occurrences")
points(allrecords$lon, allrecords$lat, col='orange', pch=20, cex=0.75)
points(allrecords$lon, allrecords$lat, col='red', cex=0.75)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
### Transform dataframe into spatial points dataframe
coordinates(world) <- ~lon+lat
crs(world) <- crs(wrld_simpl)

coordinates(europe) <- ~lon+lat
crs(europe) <- crs(wrld_simpl)

coordinates(europe_sure) <- ~lon+lat
crs(europe_sure) <- crs(wrld_simpl)
```

### Perform spatial thinning to reduce spatial autocorrelation
```{r echo=TRUE, message=FALSE, warning=FALSE}
set.seed(2018)
### To reduce sampling bias (disproportional reporting in some areas), a single observation was sampled 
# by first creating a raster layer with the same extent as my environmental data
r <- raster(world)
# set resolution of cells (increased resolution enough to obtain ~30 observations)
res(r) <- 0.026
# sample 1 observation per area
world_sampled <- gridSample(world, r, n=1)

r <- raster(europe)
res(r) <- 0.026
europe_sampled <- gridSample(europe, r, n=1)

r <- raster(europe_sure)
res(r) <- 0.026
europe_sure_sampled <- gridSample(europe_sure, r, n=1)

world_sampled <- as.data.frame(world_sampled)
europe_sampled <- as.data.frame(europe_sampled)
europe_sure_sampled <- as.data.frame(europe_sure_sampled)

plot(wrld_simpl, axes=TRUE, col="light yellow", main = "Sampled occurrences")
legend("bottomright", legend=c("Training", "Testing"),
       col=c("orange", "blue"), cex=0.8, pch = 19)
# add observation points
points(world_sampled$lon, world_sampled$lat, col='orange', pch=20, cex=0.75)
points(europe_sampled$lon, europe_sampled$lat, col='blue', pch=20, cex=0.75)
```

### Divide dataset into training (world excluding Europe) and testing (Europe)
```{r echo=TRUE, message=FALSE, warning=FALSE}
pres_train <- world_sampled
pres_test <- europe_sampled
pres_test_sure <- europe_sure_sampled
write.csv(pres_train,"pres_train.csv",row.names=FALSE)
write.csv(pres_test,"pres_test.csv",row.names=FALSE)
write.csv(pres_test_sure,"pres_test_sure.csv",row.names=FALSE)
```

### Initial selection of bioclimatic variables

- To reflect thermal dependency of duckweeds in terms of population growth:

BIO 1 = Mean annual temperature (Armitage, 2022)

BIO 7 = Annual range of temperature (Armitage, 2022)

BIO 10 = Mean temperature of warmest quarter

- To account for the dependency on precipitation (duckweeds are rare in places with high or very low precipitation; FNA, 2000):

BIO17 = Precipitation of Driest Quarter

BIO19 = Annual precipitation

### Load lake water temperature raster
```{r echo=TRUE, message=FALSE, warning=FALSE}
#source for lake water temperature raster: https://onlinelibrary.wiley.com/doi/full/10.1111/ecog.06595#bib-0003
envs_lake <- raster::brick('bioclim_lakes_10km.tif')
```

### Subset raster to correspond to the pre-selected bioclimatic variables
```{r echo=TRUE, message=FALSE, warning=FALSE}
envs = envs_lake[[c(1,7,10,17,19)]] # pre-select bioclimatic variables
```

### Increase spatial resolution to correspond to resolution previously applied to raster of observation records
```{r echo=TRUE, message=FALSE, warning=FALSE}
envs <- disaggregate(envs, fact=3.2) # adjust resolution to correspond to resolution used to sample observations
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(sp)
coordinates(world_sampled) <- ~lon + lat
coordinates(europe_sampled) <- ~lon + lat
```

### Crop environment to correspond to species distribution range
```{r message=FALSE, warning=FALSE}
full_dataset <- rbind(pres_train,pres_test)
model.extent<-extent(min(full_dataset$lon)-10,max(full_dataset$lon)+10,
                     min(full_dataset$lat)-10,max(full_dataset$lat)+10)
modelEnv <- crop(envs,model.extent)
```

### Create buffer around occurrence area and create study area 
```{r echo=TRUE, message=FALSE, warning=FALSE}
occ_buff <- buffer(world_sampled,600000, dissolve=TRUE) #width parameter=600000m = 600km

# crop study area to buffer extent
studyArea <- crop(envs,extent(occ_buff))  

# mask the non buffer areas
studyArea <- mask(studyArea,occ_buff)
# output will still be a raster stack, just of the study area

# check with plot
plot(studyArea[[1]]/10, main="Buffer areas and observation points") 
points(world_sampled$lon, world_sampled$lat, pch=1, cex=0.51, col="purple")
legend("bottomright",legend=c("Buffer areas: annual mean temperature coloring", 
                            "Occurrences: purple"))
```

### Randomly sample points 
Sample same number as our observed points inside the buffer to create pseudo-absences, or hypothetical areas where species is not found
```{r echo=TRUE, message=FALSE, warning=FALSE}
set.seed(2020)
backg = randomPoints(studyArea, n=708, p=world_sampled, extf=1)
colnames(backg) <- c("lon","lat")
```

### Divide background dataset into training and testing
```{r echo=TRUE, message=FALSE, warning=FALSE}
group=kfold(backg, 5)
backg_train <- backg[group!=1,]
backg_test <- backg[group==1,]
```

### Run model and check variable contribution
```{r echo=TRUE, message=FALSE, warning=FALSE}
xm <- maxent(modelEnv, pres_train, linear=true, quadratic=true, product=false, 
              threshold=false, hinge=false)
plot(xm)

variable <- plot(xm)
variable
```


### Check potential correlations between bioclimatic variables
```{r echo=FALSE, message=FALSE, warning=FALSE}
stats <- layerStats(modelEnv, 'pearson', na.rm=T)
corr_matrix <- stats$'pearson correlation coefficient'

row.names(corr_matrix) <- c("BIO1", "BIO7", "BIO10", "BIO17", "BIO19")
colnames(corr_matrix) <- c("BIO1", "BIO7", "BIO10", "BIO17", "BIO19")

order <- c("BIO1", "BIO7", "BIO19", "BIO17", "BIO10")
corr_matrix <- corr_matrix[order, order]
    
corr_stats <- data.frame("Preliminary variables"=c("BIO 1 (mean annual temperature)",
                                                   "BIO 7 (annual range of temperature)",
                                                   "BIO 19 (mean annual precipitation)",
                                                   "BIO17 (precipitation of driest quarter)",
                                                   "BIO 10 (mean temperature of the warmest quarter)"), 
                         "Variable contribution"=rev(variable), "BIO 1"=as.numeric(c(corr_matrix[1:1],"","","","")), 
                         "BIO 7"=as.numeric(c(corr_matrix[1:2,2],"","","")), 
                         "BIO 19"=as.numeric(c(corr_matrix[1:3,3],"","")),
                         "BIO 17"=as.numeric(c(corr_matrix[1:4,4],"")),
                         "BIO 10"=as.numeric(c(corr_matrix[1:5,5])))


opts <- options(knitr.kable.NA = "")
table <- knitr::kable(corr_stats, "html", col.names = c('Preliminary variables', 'Variable contribution', 'BIO1', 'BIO7', 'BIO19', 'BIO17', 'BIO10'), digits = 2, row.names = FALSE)
library(kableExtra)
table <- add_header_above(table, header = c(" " = 2, "Bioclimatic variables" = 5))
table
```


### Select variables which contribute the most and are unrelated to each other
* BIO 1 = Mean annual temperature
* BIO 7 = Annual range of temperature
* BIO 19 = Mean annual temperature