---
title: 'Model comparisons (rmd file #5)'
author: "Debora"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

<style type="text/css">
  body{
  font-size: 15pt;
}
</style>

```{r setup, include=FALSE}
Sys.setenv('R_MAX_VSIZE'=3200000000000)
memory.limit(24000000)

options(java.parameters="~xmx10g")
library(geodata)
library(raster)
library(dismo)
library(maps)
library(rJava)
library(ENMeval)
library(maxnet)
library(rTPC)
library(nls.multstart)
library(sp)
library(kableExtra)
library(ENMeval)
library(viridis)
library(ggplot2)
library(maps)

.jinit()

knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	include = FALSE
)
```

### Download environmental predictors
```{r message=FALSE, warning=FALSE, include=FALSE}
# Lake water temperature raster based on satellite measurements (Armitage, 2023; https://onlinelibrary.wiley.com/doi/full/10.1111/ecog.06595#bib-0003)
options(timeout=600)
url <- "https://datadryad.org/stash/downloads/file_stream/1895801"
temp_file <- tempfile()
temp_unzipped1 <- tempfile()
temp_unzipped2 <- tempfile()
download.file(url, destfile = temp_file, mode="wb")
unzip(temp_file, exdir = temp_unzipped1)
unzip(paste0(temp_unzipped1,"/LakeTemps_Code/rasters.zip"), exdir = temp_unzipped2)
lake <- raster::brick(paste0(temp_unzipped2,"/rasters/bioclim_lakes_10km.tif"))

air <- geodata::worldclim_global(var = 'bio', res = 5, download = T, path = 'data')
air <- as(air, "Raster")
air <- brick(air)
```

Future air temperature model (ACCESS)
```{r echo=TRUE, message=FALSE, warning=FALSE}
# Future climate scenario from 'ACCESS-ESM1-5' climate model
clim_fut_access <- geodata::cmip6_world(model='ACCESS-ESM1-5', ssp='245', time='2041-2060', var='bioc', download=T, res=5, path='data')
clim_fut_access <- as(clim_fut_access, "Raster")
```

Future air temperature model (MIROC)
```{r echo=TRUE, message=FALSE, warning=FALSE}
# Future climate scenario from 'MIROC-ES2L' climate model
clim_fut_miroc <- geodata::cmip6_world(model='MIROC-ES2L', ssp='245', time='2041-2060', var='bioc', download=T, res=5, path='data')
clim_fut_miroc <- as(clim_fut_miroc, "Raster")
```


```{r echo=FALSE}
### Raster alignment
# https://gist.github.com/dbaston/767e9a2d7f2e4c20c6ea1305a7351c5e

e <- as(extent(-179.8668, 180, -60.73347, 83.59986), 'SpatialPolygons')
crs(e) <- "+proj=longlat +datum=WGS84 +no_defs"
air_crop <- crop(air, e, snap="near")
lake_crop <- crop(lake, e, snap="near")
clim_fut_access_crop <- crop(clim_fut_access, e, snap="near")
clim_fut_miroc_crop <- crop(clim_fut_miroc, e, snap="near")

# round resolution to nearest arc second
snap_res <- function(r) {
  round(res(r)*3600)/3600
}

# make sure extent fits cleanly onto a grid
# having its origin at -180, -90
snap_extent <- function(r) {
  origin <- c(-180, -90)
  
  ll <- origin - round((origin - extent(r)[c(1, 3)])/res(r))*res(r)
  ur <- ll + res(r)*dim(r)[c(2, 1)]
  
  raster::extent(c(ll[1], ur[1], ll[2], ur[2]))
}


res(lake_crop) <- snap_res(lake_crop)
res(air_crop) <- snap_res(air_crop)

res(clim_fut_access_crop) <- snap_res(clim_fut_access_crop)
res(clim_fut_miroc_crop) <- snap_res(clim_fut_miroc_crop)

stopifnot(res(lake_crop) == res(air_crop))

stopifnot(res(lake_crop) == res(clim_fut_access_crop))
stopifnot(res(lake_crop) == res(clim_fut_miroc_crop))

extent(lake_crop) <- snap_extent(lake_crop)
extent(air_crop) <- snap_extent(air_crop)

extent(clim_fut_access_crop) <- snap_extent(clim_fut_miroc_crop)
extent(clim_fut_miroc_crop) <- snap_extent(clim_fut_miroc_crop)

combined_extent <- merge(extent(lake_crop), extent(air_crop))

lake_crop <- extend(lake_crop, combined_extent)
air_crop <- extend(air_crop, combined_extent)

combined_extent <- merge(extent(lake_crop), extent(clim_fut_access_crop))
clim_fut_access_crop <- extend(clim_fut_access_crop, combined_extent)

combined_extent <- merge(extent(lake_crop), extent(clim_fut_miroc_crop))
clim_fut_miroc_crop <- extend(clim_fut_miroc_crop, combined_extent)

stk <- stack(lake_crop, air_crop, clim_fut_access_crop, clim_fut_miroc_crop)

masked <- mask(x = air_crop, mask = lake_crop)
air_crop <- crop(x = masked, y = extent(lake_crop))

masked <- mask(x = clim_fut_access_crop, mask = lake_crop)
clim_fut_access_crop <- crop(x = masked, y = extent(lake_crop))

masked <- mask(x = clim_fut_miroc_crop, mask = lake_crop)
clim_fut_miroc_crop <- crop(x = masked, y = extent(lake_crop))




### Printing tip of South America to check if mask/crop worked correctly
# air1_crop_na_map <- crop(air1_crop_na,c(-100,-50,-55,-35))
# lake1_crop_map <- crop(lake1_crop,c(-100,-50,-55,-35))
# air1_crop_map_original <- crop(air1_crop,c(-100,-50,-55,-35))

# par(mfrow=c(1,3))
# plot(lake1_crop_map[[1]])
# plot(air1_crop_map_original[[1]])
# plot(air1_crop_na_map[[1]])
```

### Select bioclimatic variables
```{r}
# Showing code for a single environmental predictor source 
# (the same selection is done for all environmental rasters)
air_crop1 <- air_crop[[c(1,7,15)]]
```

```{r include=FALSE}
air_crop1 <- brick(air_crop1)

lake_crop1 <- lake_crop[[c(1,7,15)]]
lake_crop1 <- brick(lake_crop1)

air_crop11 <- air_crop[[c(10,11,15)]]
air_crop11 <- brick(air_crop11)

lake_crop11 <- lake_crop[[c(10,11,15)]]
lake_crop11 <- brick(lake_crop11)

names(clim_fut_access_crop) <- names(air_crop)
names(clim_fut_miroc_crop) <- names(air_crop)

clim_fut_access_crop1 <- clim_fut_access_crop[[c(1,7,15)]]
clim_fut_access_crop1 <- brick(clim_fut_access_crop1)

clim_fut_miroc_crop1 <- clim_fut_miroc_crop[[c(1,7,15)]]
clim_fut_miroc_crop1 <- brick(clim_fut_miroc_crop1)

clim_fut_access_crop11 <- clim_fut_access_crop[[c(10,11,15)]]
clim_fut_access_crop11 <- brick(clim_fut_access_crop11)

clim_fut_miroc_crop11 <- clim_fut_miroc_crop[[c(10,11,15)]]
clim_fut_miroc_crop11 <- brick(clim_fut_miroc_crop11)
```

### Load sampled datasets obtained in previous steps of the analysis
Training dataset (global records except Europe)
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Sampled records
pres_train <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/distribution/pres_train.csv", header=TRUE, stringsAsFactors = TRUE, fileEncoding="UTF-8-BOM")
pres_test <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/distribution/pres_test.csv", header=TRUE, stringsAsFactors = TRUE, fileEncoding="UTF-8-BOM")

head(pres_train)
```

Testing dataset (European records)
```{r echo=FALSE, message=FALSE, warning=FALSE}
head(pres_test)

# Transform dataframe into spatial points dataframe
coordinates(pres_train) <- ~lon+lat
coordinates(pres_test) <- ~lon+lat

myCRS <- CRS("+init=epsg:4326") # WGS 84
crs(pres_train) <- myCRS
crs(pres_test) <- myCRS
```


### Create study area, sample background points, and run models
```{r echo=TRUE, message=TRUE, warning=TRUE}
raster_list <- list(lake_crop1, air_crop1, lake_crop11, air_crop11)
crop_raster_list <- list()
backg_train_list <- list()
maxent_output_list <- list()

count = 1
for (i in raster_list){
# Crop environment to correspond to species distribution range
model.extent<-extent(min(pres_train$lon)-10,max(pres_train$lon)+10,
                     min(pres_train$lat)-10,max(pres_train$lat)+10)

crop_raster <- crop(i,model.extent)

crop_raster_list[[count]] <- crop_raster

occ_buff <- buffer(pres_train, 600000, dissolve=TRUE) # width parameter = 600000m (600km)

# crop study area to buffer extent
studyArea <- crop(crop_raster, extent(occ_buff))  

# mask the non buffer areas
studyArea <- mask(studyArea, occ_buff)
# output will still be a raster stack, just of the study area

# Randomly sample points 
# Sample same number as our observed points inside the buffer 
# to create pseudo-absences, or hypothetical areas where species 
# could either be found or not

set.seed(2022)
backg_train = randomPoints(studyArea, n=length(pres_train$lon), p=pres_train, extf=1)
colnames(backg_train) <- c("lon","lat")

backg_train_list[[count]] <- backg_train 

# Run candidate models
e.mx.l <- ENMevaluate(occs = pres_train, envs = crop_raster, bg = backg_train,
                                  algorithm = 'maxent.jar', partitions = 'block', 
                                  tune.args = list(fc = c("L","LQ"),             
                                  rm = seq(0.5, 4, by = 0.5)))

maxent_output_list[[count]] <- e.mx.l


count=count+1  
}

```


### Calculate model performance metrics
### Select models with lowest AICc
```{r echo=FALSE, message=FALSE, warning=FALSE}
results <- list()
results_kable <- list()
best_aic_list <- list()
best_auc_list <- list()
best_model_list <- list()
best_aic_list2 <- list()
model_predictions_list <- list()

count = 1
for (i in maxent_output_list){
  result <- eval.results(i)
  result2 <- eval.results(i)
  result <- result[,c(1,2,4,8,16,17,19)]
  colnames(result) <- c("Features","Regularization.multiplier","Training.AUC","Mean.testing.AUC","AICc","Delta AICc","Parameters") 
  result_kable <- knitr::kable(result, "simple", digits = 2)
  results[[count]] = as.data.frame(result)
  results_kable[[count]] = result_kable
  
  best_aic <- result[which.min(result$AICc),]
  best_aic2 <- result2[which.min(result2$AICc),]
  best_auc <- result[which.max(result$Mean.testing.AUC),]

  model <- eval.models(i)[[best_aic2$tune.args]]
  best_model_list[[count]] <- model
  
  best_aic_list[[count]] = as.data.frame(best_aic)
  best_auc_list[[count]] = as.data.frame(best_auc)
  best_aic_list2[[count]] = as.data.frame(best_aic2)
  
   ### Save model predictions
  pred_seq <- eval.predictions(i)[[best_aic2$tune.args]]
  model_predictions_list[[count]] <- pred_seq
  
  count=count+1  
}
```

Lake model (Bio 1,7,15)
```{r echo=FALSE, message=FALSE, warning=FALSE}
results_kable[[1]]
```

Air temperature model (Bio 1,7,15)
```{r echo=FALSE, message=FALSE, warning=FALSE}
results_kable[[2]]
```

Lake model (Bio 10,11,15)
```{r echo=FALSE, message=FALSE, warning=FALSE}
results_kable[[3]]
```

Air temperature model (Bio 10,11,15)
```{r echo=FALSE, message=FALSE, warning=FALSE}
results_kable[[4]]
```

### Obtain background points for the testing region (Europe)
Here we followed the same process as shown for the training region
```{r message=TRUE, warning=TRUE, include=FALSE}
europe_backg_list <- list()
crop_raster_europe_list <- list()

count = 1
for (i in raster_list){
  
  model.extent<-extent(min(pres_test$lon)-10,max(pres_test$lon)+10,
                     min(pres_test$lat)-10,max(pres_test$lat)+10)

  crop_raster_europe <- crop(i,model.extent)

  crop_raster_europe_list[[count]] <- crop_raster_europe

  # Crop environment to correspond to species distribution range in invaded region
  occ_buff <- buffer(pres_test, 600000, dissolve=TRUE) # width parameter = 600000m (600km)
  
  # crop study area to buffer extent
  studyArea <- crop(crop_raster_europe, extent(occ_buff))  
  
  # mask the non buffer areas
  studyArea <- mask(studyArea, occ_buff)
  
  set.seed(2024)
  europe_backg = randomPoints(studyArea, n=length(pres_test$lon), p=pres_test, extf=1)
  colnames(europe_backg) <- c("lon","lat")
  
  europe_backg_list[[count]] <- europe_backg
  
  count=count+1  
}
```

### Evaluate models
```{r echo=FALSE, message=FALSE, warning=FALSE}
metrics <- data.frame(matrix(nrow = 4, ncol = 7)) 
names(metrics) <- c("Model","AUC_train", "AUC_test", "TSS_train", "TSS_test", "Thresh_train", "Thresh_test")
metrics$Model <- c("Lake_BIO1,7,15", "Air_BIO1,7,15","Lake_BIO10,11,15", "Air_BIO10,11,15")
model_predictions_list <- list()

count=1

for (i in maxent_output_list){
  best_aic2 <- best_aic_list2[[count]]
  model_eval <- eval.models(i)[[best_aic2$tune.args]]
  crop_raster <- crop_raster_list[[count]]
  
  e_2train <- evaluate(pres_train, backg_train, model_eval, crop_raster)
  e_2test <- evaluate(pres_test, europe_backg, model_eval, crop_raster)
  
  e_2auc_train <- e_2train@auc
  e_2auc_test <- e_2test@auc
  metrics$auc_train[count] <- e_2auc_train  
  metrics$auc_test[count] <- e_2auc_test
  
  e_2tss_train <- max(e_2train@TPR + e_2train@TNR) - 1
  e_2tss_test <- max(e_2test@TPR + e_2test@TNR) - 1
  metrics$tss_train[count] <- e_2tss_train  
  metrics$tss_test[count] <- e_2tss_test
  
  #threshold value maximizing TPR + TNR
  threshold_train <- e_2train@t[which(e_2train@TPR+e_2train@TNR==max(e_2train@TPR+e_2train@TNR))]
  metrics$threshold_train[count] <- threshold_train
  threshold_test <- e_2test@t[which(e_2test@TPR+e_2test@TNR==max(e_2test@TPR+e_2test@TNR))]
  metrics$threshold_test[count] <- threshold_test

  ### Save model predictions
  pred_seq <- eval.predictions(i)[[best_aic2$tune.args]]
  model_predictions_list[[count]] <- pred_seq
  
  count=count+1 
}

knitr::kable(metrics, "simple", digits = 2)
```


### Plot model predictions for Europe
```{r message=FALSE, warning=FALSE, include=FALSE}
pres_test <- as.data.frame(pres_test)
europe_plot_list <- list()

count = 1
for (i in maxent_output_list){
  
  model_plot <- i
  best_aic2 <- best_aic_list2[[count]]
  eval_model_plot <- eval.models(model_plot)[[best_aic2$tune.args]]
  plot.extent<-extent(-20,65,31,70)
  raster_plot <- crop(raster_list[[count]],plot.extent)
  predictions_plot <- predict(raster_plot, eval_model_plot, type = 'cloglog')
  predictions_plot <- as.data.frame(predictions_plot, xy=T)

  pred_plot <- ggplot() +
  geom_raster(data = predictions_plot, aes(x = x, y = y, fill = layer)) +
  coord_map(xlim=c(-20,65),ylim=c(31,70)) +
  borders("world", xlim=c(-20,65),ylim=c(31,70)) + 
  geom_point(data=pres_test, aes(x=lon, y=lat), col='red', cex=1) +
  scale_size_area() +
  coord_quickmap(xlim=c(-20,65),ylim=c(31,70)) +
  scale_fill_gradientn(colours=rev(terrain.colors(255))) +
  theme_bw() + theme(panel.grid.major = element_blank(), 
                     panel.grid.minor = element_blank(), 
                     axis.line = element_line(colour = "black")) + 
  scale_y_continuous(expand = c(0, 0) ) +
  scale_x_continuous(expand = c(0, 0) ) +
  xlab("Longitude") + ylab("Latitude") + labs(fill = "Habitat suitability") + 
  theme(text=element_text(size=20))
  
  europe_plot_list[[count]] <- pred_plot
  
  count=count+1  
}
```

### Plot model predictions for North America
```{r message=FALSE, warning=FALSE, include=FALSE}
pres_train <- as.data.frame(pres_train)
na_plot_list <- list()

count = 1
for (i in maxent_output_list){
  
  model_plot <- i
  best_aic2 <- best_aic_list2[[count]]
  eval_model_plot <- eval.models(model_plot)[[best_aic2$tune.args]]
  plot.extent<-extent(-130,-55,21,60)
  raster_plot <- crop(raster_list[[count]],plot.extent)
  predictions_plot <- predict(raster_plot, eval_model_plot, type = 'cloglog')
  predictions_plot <- as.data.frame(predictions_plot, xy=T)
  
  pred_plot <- ggplot() +
  geom_raster(data = predictions_plot, aes(x = x, y = y, fill = layer)) +
  coord_map(xlim=c(-130,-55),ylim=c(21,60)) +
  borders("world", xlim=c(-130,-55),ylim=c(21,60)) + 
  geom_point(data=pres_train, aes(x=lon, y=lat), col='red', cex=1) +
  scale_size_area() +
  coord_quickmap(xlim=c(-130,-55),ylim=c(21,60)) +
  scale_fill_gradientn(colours=rev(terrain.colors(255))) +
  theme_bw() + theme(panel.grid.major = element_blank(), 
                     panel.grid.minor = element_blank(), 
                     axis.line = element_line(colour = "black")) + 
  scale_y_continuous(expand = c(0, 0) ) +
  scale_x_continuous(expand = c(0, 0) ) +
  xlab("Longitude") + ylab("Latitude") + labs(fill = "Habitat suitability") + 
  theme(text=element_text(size=20))
  
  na_plot_list[[count]] <- pred_plot
  
  count=count+1  
}
```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
### Checking Australia
pres_train <- as.data.frame(pres_train)
aust_plot_list <- list()

count = 1
for (i in maxent_output_list){
  
  model_plot <- i
  best_aic2 <- best_aic_list2[[count]]
  eval_model_plot <- eval.models(model_plot)[[best_aic2$tune.args]]
  plot.extent<-extent(105,160,-40,-10)
  raster_plot <- crop(raster_list[[count]],plot.extent)
  predictions_plot <- predict(raster_plot, eval_model_plot, type = 'cloglog')
  predictions_plot <- as.data.frame(predictions_plot, xy=T)
  
  pred_plot <- ggplot() +
  geom_raster(data = predictions_plot, aes(x = x, y = y, fill = layer)) +
  coord_map(xlim=c(105,160),ylim=c(-40,-10)) +
  borders("world", xlim=c(105,160),ylim=c(-40,-10)) + 
  geom_point(data=pres_train, aes(x=lon, y=lat), col='red', cex=1) +
  scale_size_area() +
  coord_quickmap(xlim=c(105,160),ylim=c(-40,-10)) +
  scale_fill_gradientn(colours=rev(terrain.colors(255))) +
  theme_bw() + theme(panel.grid.major = element_blank(), 
                     panel.grid.minor = element_blank(), 
                     axis.line = element_line(colour = "black")) + 
  scale_y_continuous(expand = c(0, 0) ) +
  scale_x_continuous(expand = c(0, 0) ) +
  xlab("Longitude") + ylab("Latitude") + labs(fill = "Habitat suitability") + 
  theme(text=element_text(size=20))
  
  aust_plot_list[[count]] <- pred_plot
  
  count=count+1  
}

aust_plot_list[[1]]
aust_plot_list[[2]]

```



```{r echo=FALSE, message=FALSE, warning=FALSE}
library(patchwork)
na_plot_list[[1]]+ theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                         legend.position = "NULL", axis.text.x=element_blank(), 
                         plot.title = element_text(size=15)) +
  ggtitle("Lake BIO 1,7,15") +
  guides(shape = guide_legend(override.aes = list(size = 0.4)),
         color = guide_legend(override.aes = list(size = 0.4))) +
  europe_plot_list[[1]]+ theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                               axis.text.x=element_blank(), 
                               plot.title = element_text(size=15),
                               legend.title = element_text(size=13), 
                               legend.text = element_text(size = 12)) +  
  na_plot_list[[2]]+ theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                           legend.position = "NULL", axis.text.x=element_blank(), 
                           plot.title = element_text(size=15)) +
  ggtitle("Air BIO 1,7,15") +
  europe_plot_list[[2]]+ theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                               legend.position = "NULL", axis.text.x=element_blank(), 
                               plot.title = element_text(size=15)) + 
  na_plot_list[[3]]+ theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                           legend.position = "NULL", axis.text.x=element_blank(), 
                           plot.title = element_text(size=15)) + 
  ggtitle("Lake BIO 10,11,15") +
  europe_plot_list[[3]]+ theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                               legend.position = "NULL", axis.text.x=element_blank(), 
                               plot.title = element_text(size=15)) + 
  na_plot_list[[4]]+theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                          plot.title = element_text(size=15))+
  theme(legend.position = "NULL") +
  ggtitle("Air BIO 10,11,15") +
  europe_plot_list[[4]]+theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                              plot.title = element_text(size=15))+
  theme(legend.position = "NULL") +  plot_layout(ncol = 2) 
```

Difference between model predictions using lake vs air temperature predictors (BIO 1,7,15)
In yellow we see areas where the 2 models differ the most (the lake model estimates more suitability)

North America
```{r echo=FALSE, message=FALSE, warning=FALSE}
a <- model_predictions_list[[1]]
b <- model_predictions_list[[2]]

difference <- a-b
# a= lake - b=air
# 0.7     - 0.4  = 0.3 (difference)

plot.extent<-extent(-20,65,31,80)
difference_europe <- crop(difference,plot.extent)
plot.extent<-extent(-130,-55,21,60)
difference_na <- crop(difference,plot.extent)
  
plot(difference_na, breaks = c(-0.25, 0.2,0.65), 
     col = c("#999999","#F0E442"),
     xlab="Longitude", ylab= "", cex.lab=1.3)
points(pres_train$lon, pres_train$lat, pch = 16, col = "red", cex = 0.5)
```

Europe
```{r echo=FALSE, message=FALSE, warning=FALSE}
plot(difference_europe, breaks = c(-0.25, 0.2,0.65), 
     col = c("#999999","#F0E442"),
     xlab="Longitude", ylab= "Latitude", cex.lab=1.3,legend = FALSE)
points(pres_test$lon, pres_test$lat, pch = 16, col = "red", cex = 0.5)
```

### Making future predictions based on best air temperature model
Future air temperature model (ACCESS)
```{r echo=FALSE, message=FALSE, warning=FALSE}
model_air_present <- maxent_output_list[[2]]
best_aic2 <- best_aic_list2[[2]]
eval_model_air_present <- eval.models(model_air_present)[[best_aic2$tune.args]]

predictions_access <- predict(clim_fut_access_crop1, eval_model_air_present, type = 'cloglog')
```

Future air temperature model (MIROC)
```{r echo=FALSE, message=FALSE, warning=FALSE}
predictions_miroc <- predict(clim_fut_miroc_crop1, eval_model_air_present, type = 'cloglog')
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
predictions_access_plot <- as.data.frame(predictions_access, xy=T)

europe_plot_access <- ggplot() +
  geom_raster(data = predictions_access_plot, aes(x = x, y = y, fill = layer)) +
  coord_map(xlim=c(-20,65),ylim=c(31,70)) +
  borders("world", xlim=c(-20,65),ylim=c(31,70)) + 
  geom_point(data=pres_test, aes(x=lon, y=lat), col='red', cex=1) +
  scale_size_area() +
  coord_quickmap(xlim=c(-20,65),ylim=c(31,70)) +
  scale_fill_gradientn(colours=viridis::viridis(99)) +
  theme_bw() + theme(panel.grid.major = element_blank(), 
                     panel.grid.minor = element_blank(), 
                     axis.line = element_line(colour = "black")) + 
  scale_y_continuous(expand = c(0, 0) ) +
  scale_x_continuous(expand = c(0, 0) ) +
  xlab("Longitude") + ylab("Latitude") + labs(fill = "Habitat suitability") + 
  theme(text=element_text(size=20))

na_plot_access <- ggplot() +
  geom_raster(data = predictions_access_plot, aes(x = x, y = y, fill = layer)) +
  coord_map(xlim=c(-130,-55),ylim=c(21,60)) +
  borders("world", xlim=c(-130,-55),ylim=c(21,60)) + 
  geom_point(data=pres_train, aes(x=lon, y=lat), col='red', cex=1) +
  scale_size_area() +
  coord_quickmap(xlim=c(-130,-55),ylim=c(21,60)) +
  scale_fill_gradientn(colours=viridis::viridis(99)) +
  theme_bw() + theme(panel.grid.major = element_blank(), 
                     panel.grid.minor = element_blank(), 
                     axis.line = element_line(colour = "black")) + 
  scale_y_continuous(expand = c(0, 0) ) +
  scale_x_continuous(expand = c(0, 0) ) +
  xlab("Longitude") + ylab("Latitude") + labs(fill = "Habitat suitability") + 
  theme(text=element_text(size=20))
```

Miroc model predictions 
  
```{r echo=FALSE, message=FALSE, warning=FALSE}
predictions_miroc_plot <- as.data.frame(predictions_miroc, xy=T)

europe_plot_miroc <- ggplot() +
  geom_raster(data = predictions_miroc_plot, aes(x = x, y = y, fill = layer)) +
  coord_map(xlim=c(-20,65),ylim=c(31,70)) +
  borders("world", xlim=c(-20,65),ylim=c(31,70)) + 
  geom_point(data=pres_test, aes(x=lon, y=lat), col='red', cex=1) +
  scale_size_area() +
  coord_quickmap(xlim=c(-20,65),ylim=c(31,70)) +
  scale_fill_gradientn(colours=viridis::viridis(99)) +
  theme_bw() + theme(panel.grid.major = element_blank(), 
                     panel.grid.minor = element_blank(), 
                     axis.line = element_line(colour = "black")) + 
  scale_y_continuous(expand = c(0, 0) ) +
  scale_x_continuous(expand = c(0, 0) ) +
  xlab("Longitude") + ylab("Latitude") + labs(fill = "Habitat suitability") + 
  theme(text=element_text(size=20))

na_plot_miroc <- ggplot() +
  geom_raster(data = predictions_miroc_plot, aes(x = x, y = y, fill = layer)) +
  coord_map(xlim=c(-130,-55),ylim=c(21,60)) +
  borders("world", xlim=c(-130,-55),ylim=c(21,60)) + 
  geom_point(data=pres_train, aes(x=lon, y=lat), col='red', cex=1) +
  scale_size_area() +
  coord_quickmap(xlim=c(-130,-55),ylim=c(21,60)) +
  scale_fill_gradientn(colours=viridis::viridis(99)) +
  theme_bw() + theme(panel.grid.major = element_blank(), 
                     panel.grid.minor = element_blank(), 
                     axis.line = element_line(colour = "black")) + 
  scale_y_continuous(expand = c(0, 0) ) +
  scale_x_continuous(expand = c(0, 0) ) +
  xlab("Longitude") + ylab("Latitude") + labs(fill = "Habitat suitability") + 
  theme(text=element_text(size=20))


na_plot_access + theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                                 legend.position = "NULL", axis.text.x=element_blank(), 
                                 plot.title = element_text(size=15)) +
  ggtitle("Access model") +
  europe_plot_access + theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                               axis.text.x=element_blank(), 
                               plot.title = element_text(size=15),
                               legend.title = element_text(size=13), 
                               legend.text = element_text(size = 12)) +  
  na_plot_miroc + theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                           legend.position = "NULL",  
                           plot.title = element_text(size=15)) +
  ggtitle("Miroc model") +
  europe_plot_miroc + theme(axis.title.x = element_blank(),axis.title.y = element_blank(), 
                               legend.position = "NULL",  
                               plot.title = element_text(size=15)) +  plot_layout(ncol = 2) 
```

# Compare TPC with response curves for BIO1
```{r message=FALSE, warning=FALSE, include=FALSE}
library(rTPC)
library(nls.multstart)

### L. minor
datin1 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2022-2023.csv",
                   header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")
datin2 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2021-2022.csv",
                   header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")
datin <- rbind(datin1,datin2)

datin$Frond_count <- datin$Frond_count_1 + datin$Frond_count_2 + datin$Frond_count_3

datin <- subset(datin, Treatment == "constant")

# remove failed experiments
datin <- subset(datin, (!Errors == "y")|is.na(Errors))

datin <- subset(datin, Species=="LP")

datin$Frond_count[datin$Frond_count == 0] <- 0.001
datin$rate <- (log(datin$Frond_count)- log(12)) / 5

datin_fit <- data.frame(temp=datin$Mean_temperature, rate=datin$rate)

### Modelling thermal performance via a Ratkowsky thermal performance curve

# choose model
mod = "ratkowsky_1983"

new_data <- data.frame(temp = seq(min(datin_fit$temp), max(datin_fit$temp), 0.5))

# get start vals
start_vals <- get_start_vals(datin_fit$temp, datin_fit$rate, model_name = mod)

# get limits
low_lims <- get_lower_lims(datin_fit$temp, datin_fit$rate, model_name = mod)
upper_lims <- get_upper_lims(datin_fit$temp, datin_fit$rate, model_name = mod)

# fit model
fit <- nls_multstart(rate~ratkowsky_1983(temp = temp, tmin, tmax, a, b),
                       data = datin_fit,
                       iter = 500,
                       start_lower = start_vals - 10,
                       start_upper = start_vals + 10,
                       lower = low_lims,
                       upper = upper_lims,
                       supp_errors = 'Y')

new_data$rates <- predict(fit, newdata = new_data)

par(mar=c(5,6,4,1)+.1)
plot(datin_fit$temp,datin_fit$rate, pch = 19, ylim = c(0,0.3),
     cex=1.3,cex.lab=1.4, cex.axis=1.4, cex.main=1.4, cex.sub=1.4, col= "#999999", 
     xlab="Temperature (C)", 
     main = "Dotted duckweed reproduction across temperatures",
     ylab=(expression(paste("Relative growth rate (",Delta, " frond day"^"-1", ")"))))
lines(new_data$temp,new_data$rate,col="#56B4E9",lwd=4)


# Reference:
# Ratkowsky, D.A., Lowry, R.K., McMeekin, T.A., Stokes, A.N., & Chandler, R.E. 
# (1983). Model for bacterial culture growth rate throughout the entire 
# biokinetic temperature range. J. Bacteriol. 154, 1222-1226.
```


```{r message=FALSE, warning=FALSE, include=FALSE}
### Plot response curves for best model
model <- best_model_list[[1]]
lake_response1 <- as.data.frame(dismo::response(model,var="bioclim_lakes_10km_1"))
colnames(lake_response1) <- c("temp","pred")
lake_response7 <- as.data.frame(dismo::response(model,var="bioclim_lakes_10km_7"))
colnames(lake_response7) <- c("temp","pred")
lake_response15 <- as.data.frame(dismo::response(model,var="bioclim_lakes_10km_15"))
colnames(lake_response15) <- c("temp","pred")

model <- best_model_list[[2]]
air_response1 <- as.data.frame(dismo::response(model,var="wc2.1_5m_bio_1"))
colnames(air_response1) <- c("temp","pred")
air_response7 <- as.data.frame(dismo::response(model,var="wc2.1_5m_bio_7"))
colnames(air_response7) <- c("temp","pred")
air_response15 <- as.data.frame(dismo::response(model,var="wc2.1_5m_bio_15"))
colnames(air_response15) <- c("temp","pred")

model <- best_model_list[[3]]
lake_response10 <- as.data.frame(dismo::response(model,var="bioclim_lakes_10km_10"))
colnames(lake_response10) <- c("temp","pred")
lake_response11 <- as.data.frame(dismo::response(model,var="bioclim_lakes_10km_11"))
colnames(lake_response11) <- c("temp","pred")
lake_response15_2 <- as.data.frame(dismo::response(model,var="bioclim_lakes_10km_15"))
colnames(lake_response15_2) <- c("temp","pred")

model <- best_model_list[[4]]
air_response10 <- as.data.frame(dismo::response(model,var="wc2.1_5m_bio_10"))
colnames(air_response10) <- c("temp","pred")
air_response11 <- as.data.frame(dismo::response(model,var="wc2.1_5m_bio_11"))
colnames(air_response11) <- c("temp","pred")
air_response15_2 <- as.data.frame(dismo::response(model,var="wc2.1_5m_bio_15"))
colnames(air_response15_2) <- c("temp","pred")
```


```{r message=FALSE, warning=FALSE, include=FALSE}
### Scale RGR values to correspond to suitability values
#define function to scale values between 0 and maximal response obtained in models
scale_values <- function(rates){(rates-min(rates))/(max(rates)-min(rates))* (max(lake_response1$pred) - 0) + 0}
new_data$scaled <- scale_values(new_data$rates)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow=c(2,3))

plot(new_data$temp,new_data$scaled, col="black",xlab="Mean temperature (\u00B0C)",ylab="Response",
     main="BIO 1",ylim=c(0,1),lwd=3,lty=4,
     xlim=c(0,33),cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5, type = "l")
lines(lake_response1$temp/10, lake_response1$pred,col="#009E73",lwd=3,lty=1)
lines(air_response1$temp, air_response1$pred,lwd=3,col="#56B4E9",lty=4)
#axis(4,cex.axis=1.5)

plot(lake_response7$temp/10, lake_response7$pred, col="#009E73",
     lwd=3, xlab="Annual variation (\u00B0C)", ylab="", main="BIO 7",lty=1,ylim=c(0,1),
     xlim=c(7,40),
     cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5, type = "l",yaxt="n")
lines(air_response7$temp, air_response7$pred, col="#56B4E9", lwd=3,lty=4,
      cex = 1.5, type = "l")

plot(lake_response15$temp, lake_response15$pred, col="#009E73",lty=1, ylim=c(0,1),xlim=c(0,100),
     lwd=3, xlab="Precipitation seasonality (CV)", main="BIO 15",ylab="", 
     cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5, type = "l",yaxt="n")
lines(air_response15$temp, air_response15$pred, col="#56B4E9", lwd=3,lty=4,
      cex = 1.5, type = "l")
legend("bottomleft", legend=c("RGR", "Lake", "Air"),
       col=c("black", "#009E73", "#56B4E9"), lty=c(3,1,4,1), cex=1, lwd=3)

plot(lake_response10$temp/10, lake_response10$pred, col="#009E73",ylim=c(0,1),
     lwd=3, xlab="Mean temp. warmest quarter (\u00B0C)",ylab="Response", main="BIO 10",lty=1,
     cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5, type = "l",
     xlim=c(5,35))
lines(air_response10$temp, air_response10$pred, col="#56B4E9", lwd=3,lty=4,
      cex = 1.5, type = "l")

plot(lake_response11$temp/10, lake_response11$pred, col="#009E73",ylim=c(0,1),
     lwd=3, xlab="Mean temp. coldest quarter (\u00B0C)", ylab="", main="BIO 11",lty=1,
     cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5, type = "l",yaxt="n",
     xlim=c(-10,30))
lines(air_response11$temp, air_response11$pred, col="#56B4E9", lwd=3,lty=4,
      cex = 1.5, type = "l")

plot(lake_response15_2$temp, lake_response15_2$pred, col="#009E73",lty=1,ylim=c(0,1),xlim=c(0,100),
     lwd=3, xlab="Precipitation seasonality (CV)", main="BIO 15",ylab="",
     cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5, type = "l", yaxt="n")
lines(air_response15_2$temp, air_response15_2$pred, col="#56B4E9", lwd=3,lty=4,
      cex = 1.5, type = "l")
```


# Get min and max climatic variable values in occurrence areas to check if response curves are realistic
```{r echo=FALSE, message=FALSE, warning=FALSE}
response_list <- list(lake_response1,lake_response7,lake_response15,
                      air_response1,air_response7,air_response15,
                      lake_response10,lake_response11,lake_response15_2,
                      air_response10,air_response11,air_response15_2)

curve_values <- data.frame(matrix(nrow = 12, ncol = 4)) 
names(curve_values) <- c("Predictor","Minimum", "Optimum", "Maximum")
curve_values$Predictor <- c("Lake_BIO1", "Lake_BIO7","Lake_BIO15",
                   "Air_BIO1", "Air_BIO7","Air_BIO15",
                   "Lake_BIO10", "Lake_BIO11","Lake_BIO15_2",
                   "Air_BIO10", "Air_BIO11","Air_BIO15_2")

count=1

for (i in response_list){
  
  response <- i
  
  if (count==1|count==2|count==7|count==8) {
  curve_values$Minimum[count] <- round(min(response$temp)/10, digits = 2)
  curve_values$Optimum[count] <- round(response$temp[which.max(response$pred)]/10, digits = 2)
  curve_values$Maximum[count] <- round(max(response$temp)/10, digits = 2)
  
  
} else {
  curve_values$Minimum[count] <- round(min(response$temp), digits = 2)
  curve_values$Optimum[count] <- round(response$temp[which.max(response$pred)], digits = 2)
  curve_values$Maximum[count] <- round(max(response$temp), digits = 2)

}
  count=count+1
}


knitr::kable(curve_values, "simple", digits = 2)

# about 5% of cells have CV>1 for BIO 15 (I am not plotting these in the response curves)
air <- crop_raster_list[[4]]
length(air[[3]][air[[3]]>100])/length(air[[3]])*100
lake <- crop_raster_list[[3]]
length(lake[[3]][lake[[3]]>100])/length(lake[[3]])*100


# checking average value of raster
x <- crop_raster_list[[3]][[2]]/10
lake.stats <- data.frame(x.mean=cellStats(x, "mean"))
x <- crop_raster_list[[1]][[3]]
air.stats <- data.frame(x.mean=cellStats(x, "mean"))
```

### Get variable importance for best models
```{r echo=FALSE, message=FALSE, warning=FALSE}
lake_model1 <- maxent_output_list[[1]]
best_aic2 <- best_aic_list2[[1]]
var.imp <- eval.variable.importance(lake_model1)[[best_aic2$tune.args]]

knitr::kable(var.imp, "simple", digits = 2, caption = "Lake model (BIO 1,7,15)")

air_model1 <- maxent_output_list[[2]]
best_aic2 <- best_aic_list2[[2]]
var.imp <- eval.variable.importance(air_model1)[[best_aic2$tune.args]]
knitr::kable(var.imp, "simple", digits = 2, caption = "Air model (BIO 1,7,15)")

lake_model11 <- maxent_output_list[[3]]
best_aic2 <- best_aic_list2[[3]]
var.imp <- eval.variable.importance(lake_model11)[[best_aic2$tune.args]]
knitr::kable(var.imp, "simple", digits = 2, caption = "Lake model (BIO 10,11,15)")

air_model11 <- maxent_output_list[[4]]
best_aic2 <- best_aic_list2[[4]]
var.imp <- eval.variable.importance(air_model11)[[best_aic2$tune.args]]
knitr::kable(var.imp, "simple", digits = 2, caption = "Air model (BIO 1,11,15)")
```
