---
title: "Autocorrelation experiment - 10 days - Surface area"
author: "Debora"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Load and Preprocess Datasets

```{r}
# Load dataset
original_dataset <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2024_10d.csv", header = TRUE, stringsAsFactors = TRUE, fileEncoding = "UTF-8-BOM")
```

```{r include=FALSE}
# Data validation and cleaning
datin <- original_dataset

# Exclude experiments outside thresholds established in methods
datin <- datin[datin$Obs_sd >= 2.1 & datin$Obs_sd <= 2.9, ]
datin <- datin[!(datin$Treatment == 0 & (datin$Obs_ac < 0 | datin$Obs_ac >= 0.29)) &
                 !(datin$Treatment == 0.95 & (datin$Obs_ac <= 0.78 | datin$Obs_ac >= 0.99)), ]
datin <- datin[datin$Obs_mean >= 36.7 & datin$Obs_mean <= 37.3, ]

# Exclude failed experiments due to contamination
datin <- subset(datin, !Errors == "y" | is.na(Errors))

# Label hot-cold and cold-hot treatments
colnames(datin)[6] <- "Label"
table(datin$Label, datin$cat_1)
levels(datin$cat_1) <- c("no autocorrelation", "hot-cold", "cold-hot")
datin$Treatment <- paste0(datin$cat_1)

# Relabel groups to simplify visualization
library(stringr)
datin$Label <- str_replace(datin$Label, "0.95", "strong autocorrelation")
datin$Label <- str_replace(datin$Label, "0", "no autocorrelation")
```

## View dataset and response variable (SA = change in surface area)
```{r echo=FALSE}
dataset <- datin
names(dataset)[names(dataset) == "Experiment_Number"] <- "Exp_run"

dataset <- within(dataset, Treatment <- relevel(factor(Treatment), ref = "no autocorrelation"))
dataset$Species <- gsub("Field_LM", "LM", dataset$Species)
dataset <- within(dataset, Species <- relevel(factor(Species), ref = "LM"))
dataset$Exp_run <- as.factor(dataset$Exp_run)

dataset$Treatment <- as.factor(dataset$Treatment)

dataset$SA <- dataset$Final_area - dataset$Initial_area

knitr::kable(head(dataset[, c(1,5,23,8,11,12,24)]))
```


## Histogram of surface areas for each species
```{r echo=FALSE}
dataset$SA <- as.numeric(dataset$SA)
par(mfrow = c(2, 3))
invisible(sapply(unique(dataset$Treatment), function(trt) {
  invisible(sapply(c("LM", "LP"), function(sp) {
    hist(dataset$SA[dataset$Species == sp & dataset$Treatment == trt],
         main = paste("Species:", sp, "\nTreatment:", trt),  
         xlab = "SA", col = "lightblue", border = "white",
         breaks = seq(min(dataset$SA, na.rm = TRUE), max(dataset$SA, na.rm = TRUE) + 0.5, by = 0.5))
  }))
}))
```

## Mixed-Effects Model Fitting

```{r echo=TRUE, message=FALSE, warning=FALSE}
results_list <- list()

for (species in unique(dataset$Species)) {
  
species_data <- subset(dataset, Species == species)
library(lme4)  

simple <- lm(SA ~ Treatment, 
                 data = species_data)

exp_number <- lmer(SA ~ Treatment + (1 | Exp_run), 
                         data = species_data)

L0 <- logLik(simple)
L1 <- logLik(exp_number)
L.ratio <- as.vector(- 2 * (L0 - L1))
p_value <- 0.5 * (1 - pchisq(L.ratio, 1))
options(scipen = 999)
print(paste("Comparing simple x mixed model:"))
cat("p_value:", p_value, "\n")

results_list[[species]] <- list(simple = simple, exp_number = exp_number)
}
```

## Model Comparison and Results

We compare the performance of the models and extract estimates, confidence intervals, and p values.
```{r echo=FALSE}
results_comparison <- list()
best_models <- list()

for (species in names(results_list)) {
  Cand.modsF <- list("no random effects" = results_list[[species]]$simple,
                     "experiment number" = results_list[[species]]$exp_number)
  
  # Get AIC values for each model using base R's AIC function
  AIC_values <- sapply(Cand.modsF, AIC)
  
  # Store the AIC values in a dataframe for easy viewing
  compareF <- data.frame(
    Name = names(AIC_values),
    AIC = AIC_values
  )
  
  # Display AIC values in the dataframe (for each species)
  results_comparison[[species]] <- knitr::kable(compareF, "simple", digits=2)
  
  # Get the model name with the lowest AIC
  best_model_name <- compareF$Name[which.min(compareF$AIC)]
  best_models[[species]] <- Cand.modsF[[best_model_name]]
  
  # Display model comparison for the current species
  print(paste("Model comparison for species:", species))
  print(results_comparison[[species]])
  cat("\n")
   # Display model summary for the current species
  print(paste("Model summary for species:", species))
  best_model <- best_models[[species]]
  print(best_model)
  cat("\n")
}

```


## Model significance testing
The best model selected in the step above is tested for significance, for each species
```{r echo=FALSE}
for (species in names(results_list)) {
  
  best_model <- best_models[[species]]
  
  # Perform ANOVA for linear model
  anova_output <- anova(best_model)
   
  # Adjust p-values for ANOVA F-statistics
  anova_output$`Pr(>F)` <- pf(anova_output$`F value`, 
                              anova_output$Df - 1, 
                              df.residual(best_model), 
                              lower.tail = FALSE)
  
  # Extract residual degrees of freedom
  residual_df <- df.residual(best_model)
  
  # Print the residual degrees of freedom
  cat("Residual degrees of freedom for species", species, ":", residual_df, "\n")
  
  # Print the model summary
  model_summary <- summary(best_model)
  
  # Print the predictor significance and ANOVA results
  print(paste("Predictor significance - species", species, ":"))
  print(anova_output)
}
```

## Visualization of Results
```{r echo=FALSE}
library(ggplot2)
library(emmeans)

prediction_list <- list()
for (species in names(results_list)) {
  
  best_model <- best_models[[species]]
  species_data <- subset(dataset, Species == species)
  
  # Use emmeans to get estimates and confidence intervals
  emmeans_results <- emmeans(best_model, ~ Treatment)
  
  # Convert emmeans output to a dataframe
  predictions <- as.data.frame(summary(emmeans_results))
  prediction_list[[species]] <- predictions
  
  # Rename columns for clarity
  predictions$Lower_CI <- predictions$lower.CL
  predictions$Upper_CI <- predictions$upper.CL
  
  # Clean treatment names
  predictions$Treatment <- gsub("Treatment", "", predictions$Treatment)
  predictions$Treatment[predictions$Treatment == "(Intercept)"] <- "no autocorrelation"
  predictions$Treatment <- factor(predictions$Treatment, levels = c("no autocorrelation", "cold-hot", "hot-cold"))
  
  # Define colors for error bars
  colors <- c("grey37", "#0072B2", "#D55E00")
  
  # Create plot
  p <- ggplot(predictions, aes(x = Treatment, y = emmean)) +
    geom_point(size = 4, color = colors) +  
    geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI, color = Treatment), width = 0.2, linewidth = 1.5) +  
    scale_color_manual(values = colors, guide = "none") +  
    labs(title = "", x = "Treatment", y = bquote("Change in surface area (cm"^2*")")) +
    theme_bw(base_size = 30) +
    theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(), 
          axis.line = element_line(colour = "black")) +
    geom_jitter(data = species_data, aes(x = Treatment, y = SA, color = Treatment), width = 0.2, size = 3, shape = 1)
  
  # Add conditional annotations
  if (species == "LM") {
    p <- p + 
      coord_cartesian(ylim = c(-2, 3.6)) +
      geom_segment(aes(x = 1, xend = 3, y = 3.5, yend = 3.5), color = "black") +
      geom_text(aes(x = 2, y = 3.6, label = "**"), size = 8) +
      geom_segment(aes(x = 1, xend = 2, y = 2.7, yend = 2.7), color = "black") +
      geom_text(aes(x = 1.5, y = 3, label = "NS"), size = 8) +
      geom_segment(aes(x = 2, xend = 3, y = 3, yend = 3), color = "black") +
      geom_text(aes(x = 2.5, y = 3.2, label = "NS"), size = 8)
  } else {
    p <- p + 
      coord_cartesian(ylim = c(-1, 2)) +
      geom_segment(aes(x = 1, xend = 3, y = 1.9, yend = 1.9), color = "black") +
      geom_text(aes(x = 2, y = 1.92, label = "*"), size = 8) +
      geom_segment(aes(x = 1, xend = 2, y = 1.6, yend = 1.6), color = "black") +
      geom_text(aes(x = 1.5, y = 1.7, label = "NS"), size = 8) +
      geom_segment(aes(x = 2, xend = 3, y = 1.7, yend = 1.7), color = "black") +
      geom_text(aes(x = 2.5, y = 1.7, label = "*"), size = 8)
  }
  
  # Print each plot within the loop
  print(p)
}

```


```{r eval=FALSE, include=FALSE}
summary <- list()
# Loop through species
for (species in names(results_list)) {
  best_model <- best_models[[species]]
  
#   # Get estimated marginal means and pairwise comparisons
#   emmeans_results <- emmeans(best_model, pairwise ~ Treatment, 
#                            lmer.df = "kenward-roger")
#   emmeans_summary <- summary(emmeans_results)
# 
#   # Get the p-value for the reference level (intercept)
#   emmeans_intercept <- as.data.frame(summary(emmeans_results, infer = c(TRUE, TRUE)))
# 
#   cat("### Model results for ", species, "\n\n")
#   
#   # Print the intercept significance
#   cat("### Intercept p-value ", species, "\n\n")
#   print(emmeans_intercept[[1]][1,])
# 
#   # View the estimated marginal means
#   cat("### Contrasts p-value ", species, "\n\n")
#   print(emmeans_summary$emmeans)
# 
#   # View pairwise comparisons with p-values
#   print(emmeans_summary$contrasts)
# 
#   cat("### Variance component of random incercept for ", species, "\n\n")
# random_effects <- VarCorr(best_model)
# 
# # Extract the random intercept variance (τ00 Exp_run)
# random_intercept_variance <- as.numeric(random_effects$Exp_run[1])
# print(sprintf("τ00 Exp_run: %.3f", random_intercept_variance))
# 
# # Extract the residual variance (σ²)
# residual_variance <- attr(random_effects, "sc")^2
# print(sprintf("σ² (Residual variance): %.3f", residual_variance))
# 
# # Calculate the Intraclass Correlation (ICC)
# # ICC = variance of random intercept / (variance of random intercept + residual variance)
# ICC <- random_intercept_variance / (random_intercept_variance + residual_variance)
# print(sprintf("ICC: %.3f", ICC))
# 
# library(performance)
# 
# # Calculate marginal and conditional R2 for the model
# r2_values <- r2_nakagawa(best_model)
# 
# # Extract and print the values with 3 decimals
# marginal_r2 <- r2_values$R2_marginal
# conditional_r2 <- r2_values$R2_conditional
# 
# cat(sprintf("Marginal R²: %.3f\n", marginal_r2))
# cat(sprintf("Conditional R²: %.3f\n", conditional_r2))

library(sjPlot)
summary[[species]] <- tab_model(best_model)
}

# View sjplot model summary
cat("Common duckweed", "\n\n")
summary[[1]]
cat("Dotted duckweed", "\n\n")
summary[[2]]
``` 



## Model diagnostics

```{r model-validation, echo=FALSE, fig.show='hold'}
for (species in names(results_list)) {
  # Get the best model for each species
  best_model <- best_models[[species]]

  if (inherits(best_model, "lmerMod")) {  # Check if the model is a mixed-effects model
    cat("Processing mixed-effects model for", species, "\n")

    # Plot residual diagnostics
    print(plot(best_model, main = paste("Residuals vs Fitted for", species)))

    # Q-Q plot for normality of residuals
    qqnorm(residuals(best_model), main = paste("Q-Q Plot for", species))
    qqline(residuals(best_model), col = "red")

    # Perform a Shapiro-Wilk test on random effects for normality
    ranef_component <- ranef(best_model)
    cat("Shapiro-Wilk test for random effects (Intercept) for", species, ":\n")
    print(shapiro.test(ranef_component[[1]][, "(Intercept)"]))

  } else if (inherits(best_model, "lm")) {  # Check if the model is a simple linear model
    cat("Processing simple linear model for", species, "\n")

    # Plot diagnostics for the best model
    par(mfrow = c(1, 4))
    print(plot(best_model, main = paste(species)))

    # Perform Shapiro-Wilk test for normality of residuals
    model_residuals <- residuals(best_model)
    shapiro_test <- shapiro.test(model_residuals)
    cat("Shapiro-Wilk test for normality of residuals for", species, ":\n")
    print(shapiro_test)

    # Perform Breusch-Pagan test for heteroscedasticity
    library(lmtest)
    bptest_result <- bptest(best_model)
    cat("Breusch-Pagan test for heteroscedasticity for", species, ":\n")
    print(bptest_result)
  } else {
    cat("Unknown model type for", species, "\n")
  }
}


# Reset plotting area
par(mfrow = c(1, 1))

# For LM:
# Residuals vs Fitted: The residuals show a slight pattern, indicating a potential mild non-linearity, and shapiro is not significant. However, deviations are small, so this may be acceptable if no strong pattern is observed in other diagnostics.
# 
# Q-Q Plot: The points mostly follow the 45-degree line, suggesting that the residuals are approximately normally distributed. This is acceptable.
# 
# Scale-Location: The spread of the residuals is relatively even across fitted values, which is acceptable and suggests homoscedasticity. Breusch-Pagan test is not significant.
# 
# Residuals vs Leverage: There are no influential outliers or leverage points that significantly deviate. This is acceptable and indicates that no single observation is unduly influencing the model.

# For LP:
# Residuals vs Fitted: Althought there is a pattern (drop in residuals at one point), shapiro is not significant.
# 
# Q-Q Plot: The residuals closely follow the theoretical line, indicating that the normality assumption is mostly satisfied. This is acceptable.
# 
# Scale-Location: Although there is a upward trend in the spread of residuals, which may indicate heteroscedasticity (non-constant variance), Breusch-Pagan test is not significant (test uses model residuals and checks for significant relationship between residuals variance and predictors).
# 
# Residuals vs Leverage: While there are some points with higher leverage, none appear to be overly influential. This is acceptable, but it’s worth monitoring high-leverage points in further analysis.
```


```{r echo=FALSE}
## Post-hoc test
results_pairwise <- list()
library(emmeans)
for (species in names(results_list)) {

  best_model <- best_models[[species]]

  pairwise <- emmeans(best_model, list(pairwise ~ Treatment), lmer.df="satterthwaite")
  pairs <- as.data.frame(pairwise$`pairwise differences of Treatment`)
  results_pairwise[[species]] <- pairs

  cat("Post-hoc Test Results for Species:", species, "\n")
  print(knitr::kable(pairs, "simple", digits=3))
  cat("\n")
}
```

## If I had used mixed linear models, these would be the results
```{r echo=FALSE}
for (species in names(results_list)) {
  # Print the predictor significance for linear mixed models
  
  print(paste("Anova - linear mixed model - species", species, ":"))
  mx_anova <- car::Anova(results_list[[species]]$exp_number, type = "II")
  print(mx_anova)
  
  print(paste("Summary - linear mixed model - species", species, ":"))
  mx_summary <- summary(results_list[[species]]$exp_number)
  print(mx_summary)
  
  pairwise <- emmeans(results_list[[species]]$exp_number, list(pairwise ~ Treatment), lmer.df="satterthwaite")
  pairs <- as.data.frame(pairwise$`pairwise differences of Treatment`)
  results_pairwise[[species]] <- pairs

  cat("Post-hoc Test Results for Species:", species, "\n")
  print(knitr::kable(pairs, "simple", digits=3))
  cat("\n")
}
```

