---
title: "Autocorrelation experiment - 5 days - Mortality"
author: "Debora"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Load and Preprocess Datasets

We load two datasets: one with a preparation technique (repeated frond selection) and one without.

```{r}
options(scipen=999)
# This dataset contains replicates for which a preparation technique was performed (repeated first born frond selection to reduce maternal effects)
original_dataset_2 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2022-2023.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")

# This dataset contains replicates of experiments performed without a preparation technique 
original_dataset_1 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2021-2022.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")
```

## Combine and clean up datasets

We combine both datasets and filter out rows based on specific conditions for standard deviation (`Obs_sd`) and autocorrelation (`Obs_ac`) to clean the data.

```{r}
# Blending both datasets, as there are no significant differences between preparation methods (probably because we only have 2 methods in low temperatures)
datin <- rbind(original_dataset_1,original_dataset_2)

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9))
               &!(datin$Treatment == 0.95 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9)),]

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_ac <= -0.2 | datin$Obs_ac >= 0.2))
               &!(datin$Treatment == 0.95 & (datin$Obs_ac <= 0.92 | datin$Obs_ac >= 0.98)),]

datin <- subset(datin, !Errors == "y"|is.na(Errors))
datin <- subset(datin, !Treatment == "constant")
```


```{r include=FALSE}
## Data Relabeling
# We relabel the groups for easier visualization and manipulate the dataset to  prepare it for analysis.

colnames(datin)[6] <- "Label"
table(datin$Label, datin$cat_1)
levels(datin$cat_1) = c("no autocorrelation","hot-cold","no autocorrelation","cold-hot")
datin$Treatment<-paste0(datin$cat_1)

# Relabel groups to further simplify visualization of results
library(stringr)
datin$Label <- str_replace(datin$Label, "0.95", "strong autocorrelation")
datin$Label <- str_replace(datin$Label, "0", "no autocorrelation")

# Extract experiment run IDs
datin$Exp_run <- str_sub(datin$Profile_name,-2,-1)
datin$Exp_run <- gsub('_','',datin$Exp_run)
datin$Exp_run <- as.numeric(datin$Exp_run)
```


```{r include=FALSE}
## Subset Data
# We subset the data to retain only those experiment runs that have at least 3  replicates within each `Mean_Temperature` and `Species` level

 datin <- datin[
   ave(seq_along(datin$Exp_run), 
       interaction(datin$Exp_run, datin$Mean_temperature, datin$Species), 
       FUN = length) >= 3, 
 ]
```

```{r include=FALSE}
## Further Data Processing

# Blending 10C and 15C, as there are no significant differences between performance in these 2 average temperatures
datin$Mean_temperature <- str_replace(datin$Mean_temperature, "15", "10")
datin$Mean_temperature <- str_replace(datin$Mean_temperature, "10", "10-15")

# Removing "lab" strain, as preliminary statistical analyses showed significant differences between the 2 common duckweed strains used in this experiment in average temperature of 27C
datin <- subset(datin, !(Species == "Lab_LM" & Mean_temperature == "27"))

# Blending common duckweed strains, as preliminary analyses showed no differences between them
datin$Species <- gsub("Lab_LM|Field_LM", "LM", datin$Species)

table(datin$Mean_temperature,datin$Treatment)
```

## Dataset: each row of the dataset corresponds to a replicate; here we are showing 3 experiment runs, each including 3 treatments
```{r echo=FALSE}
datins <- subset(datin, Mean_temperature=="37")

datins$Mortality <- datins$Dead_frond_count_1 + datins$Dead_frond_count_2 + datins$Dead_frond_count_3

datins$Frond_count <- datins$Frond_count_1 + datins$Frond_count_2 + datins$Frond_count_3

datins$Proportion <- (datins$Frond_count/(datins$Frond_count+datins$Mortality))
datins$Total_frond <- datins$Frond_count+datins$Mortality

# Convert Treatment to factor and set levels
datins$Treatment <- as.factor(datins$Treatment)
datins <- within(datins, Treatment <- relevel(Treatment, ref = "no autocorrelation"))

datins[, 36] <- round(datins[, 36], 2)
knitr::kable(head(datins[, c(33, 32, 8, 35, 34, 37)]))
```

## Overview of data distribution
```{r echo=FALSE, message=FALSE, warning=FALSE}
datins$Mortality <- as.numeric(datins$Mortality)
par(mfrow = c(2, 3))
invisible(sapply(unique(datins$Treatment), function(trt) {
  invisible(sapply(c("LM", "LP"), function(sp) {
    hist(datins$Mortality[datins$Species == sp & datins$Treatment == trt],
         main = paste("Species:", sp, "\nTreatment:", trt),  
         xlab = "Mortality", col = "lightblue", border = "white",
         breaks = seq(0, max(datins$Mortality, na.rm = TRUE) + 5, by = 5))
  }))
}))
```

## Summary table: total sum of mortality across treatments
```{r echo=FALSE}
# Aggregate total dead frond count by Treatment and Species
table <- aggregate(Mortality ~ Treatment + Species, data = datins, sum)

# Rename Species
table$Species <- ifelse(table$Species == "LM", "Common duckweed", "Dotted duckweed")

# Reshape the data to have Species as columns
result_table <- as.data.frame.matrix(xtabs(Mortality ~ Treatment + Species, data = table))

print(knitr::kable(result_table,col.names = c("Treatment", "Common duckweed", "Dotted duckweed"), align = c('l', 'c', 'c'),format = "simple"))

```

## Model Fitting
```{r echo=TRUE, message=FALSE, warning=FALSE}
results_list <- list()
compare_results <- list()

for (species in unique(datins$Species)) {

  species_data <- subset(datins, Species == species)

  simple <- glm(Mortality ~ Treatment, data = species_data, family = poisson)
  library(lme4)
exp_number <- glmer(Mortality ~ Treatment + (1|Exp_run), data=species_data, family=poisson, control = glmerControl(optimizer = "nloptwrap"))

# Store models
results_list[[species]] <- list(simple = simple, exp_number = exp_number)

# Mixed models were selected for all analyses, as likelihood-ratio tests indicated that the random effects were significant in most cases
results_list[[species]] <- exp_number

  best_model <- results_list[[species]]
  
  model_dispersion <- sum(residuals(best_model, type = "pearson")^2) / df.residual(best_model)
  print(paste("Model dispersion ratio for species", species, ":", model_dispersion))
  
  cat("Wald test type 2 for significance of predictor:", species, "\n")
  library(car)
phi <- sum(residuals(best_model, type="pearson")^2) / df.residual(best_model)  
anova_table <- Anova(best_model, type = "II")  

# Adjusting chi-square values (quasipoisson)
anova_table$`Chisq` <- anova_table$`Chisq` / phi  

# Recalculating p-values using chi-square distribution
anova_table$`Pr(>Chisq)` <- pchisq(anova_table$`Chisq`, anova_table$Df, lower.tail = FALSE)  

  print(paste("Adjusted predictor significance", species, ":"))
print(anova_table)
}
```




```{r echo=FALSE}
for (species in unique(datins$Species)) {
  if (species == "LM") {
  full_mod1 <- results_list[[species]]
quasi_table <- function(model,ctab=coef(summary(model))) {
  phi <- sum(residuals(model, type="pearson")^2)/df.residual(model)
  qctab <- within(as.data.frame(ctab),
                  {`Std. Error` <- `Std. Error`*sqrt(phi)
                  `z value` <- Estimate/`Std. Error`
                  `Pr(>|z|)` <- 2*pnorm(abs(`z value`), lower.tail=FALSE)
                  })
  return(qctab)
}
print(paste("Original model summary - species :", species))
print(summary(full_mod1))

print(paste("Adjusted model summary - species :", species))
printCoefmat(quasi_table(full_mod1),digits=2)
  }
}
```  

## Post-hoc test for common duckweed
```{r echo=FALSE}
library(emmeans)

results_pairwise <- list()

for (species in names(results_list)) {
  if (species == "LM") {
  best_model <- results_list[[species]]
  
  # Overdispersion factor (phi)
  phi <- sum(residuals(best_model, type = "pearson")^2)/df.residual(best_model)  # Use nobs() for number of observations
  
  # Perform pairwise comparisons
  pairwise <- emmeans(best_model, pairwise ~ Treatment, type = "response")
  
  pairwise_adj <- emmeans(best_model, pairwise ~ Treatment, type = "response", vcov. = vcov(best_model)*phi)
  #https://stackoverflow.com/questions/78357281/post-hoc-comparisons-of-quasi-family-glmer-models-with-emmeans
  # Store results in the list
  results_pairwise[[species]] <- pairwise_adj
  
  # Print results for the species
  cat("Non-adjusted post-hoc for species:", species, "\n")
  print(pairwise)
  cat("\n")
      
  cat("Adjusted post-hoc for species:", species, "\n")
  print(pairwise_adj)
  cat("\n")
}}
```

