---
title: "Autocorrelation_stats"
author: "Debora"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Load and Preprocess Datasets

We load two datasets: one with a preparation technique (repeated frond selection) and one without.

```{r}
# This dataset contains replicates for which a preparation technique was performed (repeated first born frond selection to reduce maternal effects)
original_dataset_2 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2022-2023.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")
original_dataset_2$prep <- rep("yes",times=length(original_dataset_2$Experiment_Number))

# This dataset contains replicates of experiments performed without a preparation technique 
original_dataset_1 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2021-2022.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")
original_dataset_1$prep <- rep("no",times=length(original_dataset_1$Experiment_Number))
```

## Combine Datasets

We combine both datasets and filter out rows based on specific conditions for standard deviation (`Obs_sd`) and autocorrelation (`Obs_ac`) to clean the data.

```{r}
# Blending both datasets, as there are no significant differences between preparation methods (probably because we only have 2 methods in low temperatures)
datin <- rbind(original_dataset_1,original_dataset_2)

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9))
               &!(datin$Treatment == 0.95 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9)),]

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_ac <= -0.2 | datin$Obs_ac >= 0.2))
               &!(datin$Treatment == 0.95 & (datin$Obs_ac <= 0.92 | datin$Obs_ac >= 0.98)),]

datin <- subset(datin, !Errors == "y"|is.na(Errors))
datin <- subset(datin, !Treatment == "constant")
```


```{r include=FALSE}
## Data Relabeling
# We relabel the groups for easier visualization and manipulate the dataset to  prepare it for analysis.

colnames(datin)[6] <- "Label"
table(datin$Label, datin$cat_1)
levels(datin$cat_1) = c("no autocorrelation","hot-cold","no autocorrelation","cold-hot")
datin$Treatment<-paste0(datin$cat_1)

# Relabel groups to further simplify visualization of results
library(stringr)
datin$Label <- str_replace(datin$Label, "0.95", "strong autocorrelation")
datin$Label <- str_replace(datin$Label, "0", "no autocorrelation")

datin$Exp_run <- paste0(datin$Mean_temperature, str_sub(datin$Profile_name, -2, -1))
datin$Exp_run <- gsub('_', '', datin$Exp_run)
datin$Exp_run <- as.numeric(datin$Exp_run)
```


```{r include=FALSE}
## Subset Data
# We subset the data to retain only those experiment runs that have at least 3  replicates within each `Mean_Temperature` and `Species` level

datin <- datin[
  ave(seq_along(datin$Exp_run), 
      interaction(datin$Exp_run, datin$Mean_temperature, datin$Species), 
      FUN = length) >= 3, 
]
```

```{r include=FALSE}
## Further Data Processing

# Blending 10C and 15C, as there are no significant differences between performance in these 2 average temperatures
datin$Mean_temperature <- str_replace(datin$Mean_temperature, "15", "10")
datin$Mean_temperature <- str_replace(datin$Mean_temperature, "10", "10-15")

# Removing "lab" strain, as we did not separate generations for them and this had a significant effect at this mean temperature
datin <- subset(datin, !(Species == "Lab_LM" & Mean_temperature == "27"))

# Blending common duckweed strains, as preliminary analyses showed no differences between them
datin$Species <- gsub("Lab_LM|Field_LM", "LM", datin$Species)
```


## View dataset and response variable
```{r echo=FALSE}
datin$change_frond_count <- (datin$Frond_count_1 + datin$Frond_count_2 + datin$Frond_count_3)-12
datin$total_living_fronds <- (datin$Frond_count_1 + datin$Frond_count_2 + datin$Frond_count_3)
knitr::kable(tail(datin[, c(8, 34, 33, 36)]))

dataset <- datin
dataset <- within(dataset, Mean_temperature <- relevel(factor(Mean_temperature), ref = "10-15"))
dataset <- within(dataset, Treatment <- relevel(factor(Treatment), ref = "no autocorrelation"))
dataset <- within(dataset, Species <- relevel(factor(Species), ref = "LM"))

dataset$Experiment_Number <- as.factor(dataset$Experiment_Number)
dataset$Incubator <- as.factor(dataset$Incubator)
dataset$Mean_temperature <- as.factor(dataset$Mean_temperature)
dataset$Treatment <- as.factor(dataset$Treatment)
```



```{r echo=FALSE}
dataset$total_living_fronds <- as.numeric(dataset$total_living_fronds)
# For each species
invisible(sapply(c("LM", "LP"), function(sp) {
  
  # Create a 3x3 grid for plotting (3 treatments x 3 mean temperatures)
  par(mfrow = c(3, 3))
  
  # Set the desired order of mean temperatures
  mean_temp_order <- c("10-15", "27", "37")
  
  # Loop through each combination of treatment and mean temperature (in specified order)
  invisible(sapply(unique(dataset$Treatment), function(trt) {
    invisible(sapply(mean_temp_order, function(temp) {
      
      # Plot histogram for each combination of species, treatment, and mean temperature
      hist(dataset$total_living_fronds[dataset$Species == sp & dataset$Treatment == trt & dataset$Mean_temperature == temp],
           main = paste("Species:", sp, "\nTreatment:", trt, "\nTemperature:", temp),  
           xlab = "Live fronds", col = "lightblue", border = "white",
           breaks = seq(0, max(dataset$total_living_fronds, na.rm = TRUE) + 10, by = 10))
    }))
  }))
  
}))


```

## View number of replicates
LM
```{r echo=FALSE}
table(dataset$Mean_temperature[dataset$Species == "LM"], dataset$Treatment[dataset$Species == "LM"])
```

LP
```{r echo=FALSE}
table(dataset$Mean_temperature[dataset$Species == "LP"], dataset$Treatment[dataset$Species == "LP"])
```

## Mixed-Effects Model Fitting

We fit candidate simple or mixed-effects models to the data, including random effects for Experiment run (`Exp_run`).

```{r echo=FALSE}
species_list <- c("LM", "LP")
model_results <- list()

for (sp in species_list) {
  dataset_species <- subset(dataset, Species == sp)
  
  # Assigning new IDs to experiment runs to ensure IDs are unique (as different
  # average temperatures had same IDs)
  dataset_species$Exp_run <- rep(seq(1, length(dataset_species$Exp_run) / 3), each = 3)

  # Fit mixed-effects models for each species
  library(lme4)
  simple <- glm(total_living_fronds ~ Treatment*Mean_temperature, data=dataset_species, family=poisson)
  exp_number <- glmer(total_living_fronds ~ Treatment*Mean_temperature + (1|Exp_run), data=dataset_species, family=poisson, control = glmerControl(optimizer = "nloptwrap"))

#anova(exp_number, simple, test = "Chisq")
L0 <- logLik(simple)
L1 <- logLik(exp_number)
L.ratio <- as.vector(- 2 * (L0 - L1))
p_value <- 0.5 * (1 - pchisq(L.ratio, 1))
options(scipen = 999)
print(paste("Comparing simple x mixed model:"))
cat("p_value:", p_value, "\n")

  model_results[[sp]] <- list(simple = simple, exp_number = exp_number)
}
```

## Model Comparison and Results

We compare the performance of the models and extract estimates, confidence intervals, and p values.
```{r echo=FALSE}
# Initialize lists to store results
compare_results <- list()
best_models <- list()

# Loop over species in the list
for (sp in species_list) {
  
  # Get the models from Cand.modsF for the species
  Cand.modsF <- list("no random effects" = model_results[[sp]]$simple,
                     "experiment number" = model_results[[sp]]$exp_number)

  # Get AIC values for each model using base R's AIC function
  AIC_values <- sapply(Cand.modsF, AIC)

  # Store the AIC values in a dataframe for easy viewing
  compareF <- data.frame(
    Name = names(AIC_values),
    AIC = AIC_values
  )

  # Display the AIC values in a table using knitr
  compare_results[[sp]] <- knitr::kable(compareF, "simple", digits = 2)

  # Find the name of the model with the lowest AIC
  best_model_name <- compareF$Name[which.min(compareF$AIC)]

  # Assign the best model based on the AIC value
  best_models[[sp]] <- Cand.modsF[[best_model_name]]
  
  # Print model comparison for the current species
  print(paste("Model comparison for species:", sp))
  print(compare_results[[sp]])
}

```

## Model significance testing
```{r echo=FALSE}
for (sp in species_list) {
  best_model <- best_models[[sp]]
  
  cat("Wald test type 3 for significance of predictor:", sp, "\n")
  library(car)
  print(Anova(best_model, type = "III"))
  
  # Get the residual degrees of freedom
  residual_df <- df.residual(best_model)
  
  print(paste("Residual degrees of freedom - species :", sp, ":", residual_df))
  
  }
```

## R2 for common duckweeds
```{r echo=FALSE}
library(sjPlot)

best_model <- best_models[[1]]

# Capture the printed output of tab_model as text
summary_text <- capture.output(tab_model(best_model))

# Find the line containing "Marginal R2 / Conditional R2"
r2_line <- summary_text[grep("Marginal R2 / Conditional R2", summary_text)]

# Extract the R2 numbers using regex
library(stringr)
r2_values <- str_match(r2_line, "(\\d+\\.\\d+) / (\\d+\\.\\d+)")[, 2:3]

# Convert to numeric
marginal_r2 <- as.numeric(r2_values[1])
conditional_r2 <- as.numeric(r2_values[2])

cat("Marginal R2:", marginal_r2, "\n")
cat("Conditional R2:", conditional_r2, "\n")

```

## Visualization of Results for common duckweeds

```{r echo=FALSE}
  results_model <- summary(best_models[[1]])
  print(results_model)
```

## Model validation for common duckweeds

```{r echo=FALSE}
  model_dispersion <- sum(residuals(model_results[[1]]$exp_number, type = "pearson")^2) / df.residual(model_results[[1]]$exp_number)
  print(paste("Model dispersion ratio", ":", model_dispersion))

  # Shapiro test for random effects normality
  r_int <- ranef(model_results[[1]]$exp_number)$Exp_run$`(Intercept)`
  shapiro <- shapiro.test(r_int)
  print(paste("Shapiro test for normality of random factors", round(shapiro$p.value, 3)))
```

## Post-hoc test for common duckweeds
```{r echo=FALSE}
library(lme4)
  library(emmeans)
  best_model <- best_models[[1]]
  
  posthoc <- emmeans(best_model, pairwise ~ Treatment | Mean_temperature, type = "response")
  
  emmeans_data <- as.data.frame(posthoc$emmeans)  

  print(paste("Pairwise interactions"))
  print(emmeans_data)
  print(posthoc$contrasts)
```

```{r echo=FALSE}
library(ggplot2)
dataset_species <- subset(dataset, Species == "LM")

summary_data <- aggregate(total_living_fronds ~ Treatment + Mean_temperature, data = dataset_species, FUN = function(x) c(mean = mean(x, na.rm = TRUE), se = sqrt(mean(x) / length(x))))
summary_data <- do.call(data.frame, summary_data)

plot <- ggplot(summary_data, aes(x = Mean_temperature, y = total_living_fronds.mean, 
                                 color = Treatment, group = Treatment)) +
  geom_errorbar(aes(ymin = total_living_fronds.mean - total_living_fronds.se, ymax = total_living_fronds.mean + total_living_fronds.se), 
                width = 0.3, linewidth = 1.5, position = position_dodge(width = 0.8)) +  
  geom_point(size = 4, position = position_dodge(width = 0.8)) +  
  scale_color_manual(values = c("no autocorrelation" = "grey37", 
                                "cold-hot" = "#0072B2", 
                                "hot-cold" = "#D55E00")) +
  theme_bw(base_size = 24) +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        legend.position = "none",
        axis.line = element_line(colour = "black")) +
  labs(y = "Number of live fronds", x = expression("Mean temperature (" * degree * "C)")) +  
  guides(color = guide_legend(title = "Treatment"), linetype = guide_legend(title = "Treatment")) +
  ylim(10, 90) +
  
  geom_text(aes(x = 3.25, y = 35, label = "*"), color = "black", vjust = -0.5, size = 7) +
  
  geom_segment(aes(x = 3.05, y = 38, xend = 2.65, yend = 38), color = "black", size = 1) +
  geom_segment(aes(x = 3.3, y = 38, xend = 3.2, yend = 38), color = "black", size = 1)

plot

```