---
title: "Autocorrelation experiment: statistical analyses at 37°C - dry weight"
author: "Debora"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment = NA)
```

```{r, include=F, warning=FALSE, message=FALSE}
datin1 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/metafile_autocorrelation_2022-2023.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")

datin2 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/metafile_autocorrelation_2021-2022.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")

datin <- rbind(datin1,datin2)

# Data validation: excluding experiments outside thresholds established in my methods section
# (for standard deviation and autocorrelation of observed temperatures)
datin <- datin[!(datin$Treatment == 0 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9))
              &!(datin$Treatment == 0.95 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9)),]

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_ac <= -0.2 | datin$Obs_ac >= 0.2))
               &!(datin$Treatment == 0.95 & (datin$Obs_ac <= 0.92 | datin$Obs_ac >= 0.98)),]

# Excluding failed experiments
datin <- datin[!datin$Experiment_Number == 83,]
datin <- datin[!(datin$Experiment_Number == 99 & datin$Incubator ==3),]
datin <- datin[!(datin$Experiment_Number == 99 & datin$Incubator ==4),]

# Selecting average temperature of 37C
datin <- subset(datin, Mean_temperature == 37)

# Create new treatment label (depending on slope of regression for observed temperature series)
table(datin$Treatment, datin$cat_1)
levels(datin$cat_1) = c("N/A","N","P","N/A")
datin$label<-paste0(datin$Treatment, datin$cat_1)
table(datin$label)

# Re-label groups to simplify visualization of results
library(stringr)
datin$label <- str_replace(datin$label, "0N/A", "0")
datin$label <- str_replace(datin$label, "0.95N", "hot-cold")
datin$label <- str_replace(datin$label, "0.95P", "cold-hot")

datin$Species <- str_replace(datin$Species, "Field_LM", "L. minor (native)")
datin$Species <- str_replace(datin$Species, "LP", "L. punctata (invasive)")

datin$Species <- as.factor(datin$Species)
datin$label <- as.factor(datin$label)

# Exclude zero data and obtain final weight within each replicate (final-initial)
datinW <- subset(datin, !Initial_weight == 0)
datinW$Weight <- datinW$Final_weight - datinW$Initial_weight
datinW <- subset(datinW, !Weight == 0)
datinW <- subset(datinW, !Weight < 0)
```

### Check dataset (each row corresponds to a replicate)
```{r,echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(tail(datinW[c(1,5,7,8,32,33)],6),"simple")
```

### Check number of replicates
```{r, echo=FALSE, message=FALSE, warning=FALSE}
table(datinW$Species,datinW$label)
```

### Barplot (means and standard errors): duckweed growth 
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(cowplot)
library(rlang)
library(patchwork)

datunique.mean <- aggregate(Weight ~ label + Species, data = datinW, mean)
datunique.sd <- aggregate(Weight ~ label + Species, data = datinW, sd)
datunique <- data.frame(cbind(label=as.character(datunique.mean$label),
                              Species=as.character(datunique.mean$Species),
                              Mean=datunique.mean$Weight,
                              sd=datunique.sd$Weight))
datunique$Mean <- as.numeric(datunique$Mean) 
datunique$sd <- as.numeric(datunique$sd)

w <- ggplot(datunique, aes(x=label, y=Mean, fill=Species)) +
  geom_bar(stat = "identity",
  position = position_dodge()) +
  geom_errorbar(aes(x=label, ymin=Mean-sd, ymax=Mean+sd), width=0.4, 
                colour="grey", alpha=0.9, size=1.3, stat = "identity",
                position=position_dodge(.9)) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),
        text = element_text(size = 18),legend.position="none") +
  labs(y= "Mean dry weight", x = "Treatment") + 
  scale_fill_manual(values=c("#999999", "#E69F00"))

ggp1_legend <- ggplot(datunique, aes(x=label, y=Mean, fill=Species)) +
  geom_bar(stat = "identity",
           position = position_dodge()) + 
  theme(legend.position = "bottom")
extract_legend <- function(my_ggp) {
  step1 <- ggplot_gtable(ggplot_build(my_ggp))
  step2 <- which(sapply(step1$grobs, function(x) x$name) == "guide-box")
  step3 <- step1$grobs[[step2]]
  return(step3)
}

ggp1_legend1 <- ggp1_legend + scale_fill_manual(values = c("#999999", "#E69F00"))

shared_legend <- extract_legend(ggp1_legend1)

library("gridExtra")
grid.arrange(arrangeGrob(w, ncol = 1),
             shared_legend, nrow = 2, heights = c(10, 1))
```


### Checking the probability distribution of the response variables and removing outliers
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(MASS)
library(car)
qqp(datinW$Weight, "norm")

boxplot(datinW$Weight, plot=FALSE)$out
outliers <- boxplot(datinW$Weight, plot=FALSE)$out
datinW1 <- datinW
datinW1 <- datinW1[-which(datinW1$Weight %in% outliers),]

qqp(datinW1$Weight, "norm")
```

### Testing several models
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(lme4)

lm.fit.spW = lm(Weight ~ label*Species,  data=datinW1)
experimW <- lmer(Weight ~ label*Species + (1|Experiment_Number),  data=datinW1)
incubW <- lmer(Weight ~ label*Species + (1|Incubator),  data=datinW1)
inc_expW <- lmer(Weight ~ label*Species + (1|Incubator)+(1|Experiment_Number), data=datinW1)
exp.incW <- lmer(Weight ~ label*Species + (1|Experiment_Number/Incubator), data=datinW1)

Cand.modsW <- list("no random effects" = lm.fit.spW,
                  "experiment number" = experimW,
                  "incubator" = incubW, 
                  "incubator and experiment number" = inc_expW,
                  "incubator nested in experiment number" = exp.incW)
```

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(performance)
compareW <- compare_performance(Cand.modsW)
knitr::kable(compareW[,c(1,2,3,4)],"simple")
```

### The best model
```{r echo=TRUE, message=FALSE, warning=FALSE}
exp.incW <- lmer(Weight ~ label*Species + (1|Experiment_Number/Incubator), data=datinW1)
```

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(sjPlot)
tab_model(exp.incW, digits = 3)
```

### Dry weight analysis summary: 
- removed from overall analysis because only one of the results is significant (high risk of type 2 error) when including outliers; without outliers, no significant results were found
- residuals deviate from normality (left-skewed), log-transforming data did not resolve this issue