---
title: 'Anova for the pea-aphid experiment at 15&deg;C - autocorrelated temperatures '
author: "Kim: based on tidyverse dependent code from Jon, Champ, Sunandan, Debora"
date: "10/08/2021"
output: github_document
always_allow_html: true
---
### Purpose
To determine if there is a significant relationship between aphid offspring (dependent variable) and different autocorrelated temperature regimes (independent variable)

### Download the data from github repo and check import
```{r message=FALSE, warning=FALSE}
datin <- read.csv("https://raw.githubusercontent.com//Cuddington-Lab/thermal-experiments/main/expdata_metafile_01062021.csv",
              header=TRUE, stringsAsFactors = TRUE)
str(datin)
```
### Complete any data manipulations

Create new data frame including only mean temp of 15&deg;C, remove 0.9 autocorrelation treatments, and check
 
```{r}
dat <- datin[datin$Mean_Temp == 15 & datin$Autocorrelation!=0.9,]
table(dat$Mean_Temp,dat$Autocorrelation)
```

Create new treatment label and check
```{r}
table(dat$Autocorrelation, dat$Cat)
levels(dat$Cat) = c("m<0","", "m>0" )
dat$label<-paste0(dat$Autocorrelation, dat$Cat)
table(dat$label)
```
So we have 2, 0.95 treatments without a slope estimate??

Create new column including sum of offspring (sumOff) 
```{r}
str(dat)
dat$sumOff=dat$Offspring_Plant1+dat$Offspring_Plant2+dat$Offspring_Plant3
```

### Plot 
```{r box, fig.cap="Boxplot of mean=15C treatments"}

tr=boxplot(sumOff~label, data=dat,main=expression(paste("mean temperature 15",degree,"C")),outline=FALSE,
           xlab="autocorr treatment", ylab="summed offspring",
  names = levels(as.factor(dat$label)))

stripchart(sumOff~label, data=dat, 
           vertical = TRUE, method = "jitter",
           pch = 19, add = TRUE) 
         
```


### Stats

```{r}
 
 l1=lm(sumOff~label, data=dat)
 summary(l1)
```

### Format results

Adding a table of results for Jorren's benefit

```{r}
 library(kableExtra)
kable_classic(kable(summary(l1)$coefficients,caption="Anova results for mean 15&deg;C treatments"))

```
   
