---
title: "Autocorrelation_stats"
author: "Debora"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Load and Preprocess Datasets

We load two datasets: one with a preparation technique (repeated frond selection) and one without.

```{r}
library(interactions)
library(ggplot2)
  
# This dataset contains replicates for which a preparation technique was performed (repeated first born frond selection to reduce maternal effects)
original_dataset_2 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2022-2023.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")
original_dataset_2$prep <- rep("yes",times=length(original_dataset_2$Experiment_Number))

# This dataset contains replicates of experiments performed without a preparation technique 
original_dataset_1 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/dataset_autocorrelation_2021-2022.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")
original_dataset_1$prep <- rep("no",times=length(original_dataset_1$Experiment_Number))
```

## Combine datasets and do data validation

We combine both datasets and filter out rows based on specific conditions for standard deviation (`Obs_sd`) and autocorrelation (`Obs_ac`) to clean the data.

```{r}
# Blending both datasets, as there are no significant differences between preparation methods (probably because we only have 2 methods in low temperatures)
datin <- rbind(original_dataset_1,original_dataset_2)

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9))
               &!(datin$Treatment == 0.95 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9)),]

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_ac <= -0.2 | datin$Obs_ac >= 0.2))
               &!(datin$Treatment == 0.95 & (datin$Obs_ac <= 0.92 | datin$Obs_ac >= 0.98)),]

datin <- subset(datin, !Errors == "y"|is.na(Errors))
datin <- subset(datin, !Treatment == "constant")
```


```{r include=FALSE}
## Data Relabeling
# We relabel the groups for easier visualization and manipulate the dataset to  prepare it for analysis.

colnames(datin)[6] <- "Label"
table(datin$Label, datin$cat_1)
levels(datin$cat_1) = c("no autocorrelation","hot-cold","no autocorrelation","cold-hot")
datin$Treatment<-paste0(datin$cat_1)

# Relabel groups to further simplify visualization of results
library(stringr)
datin$Label <- str_replace(datin$Label, "0.95", "strong autocorrelation")
datin$Label <- str_replace(datin$Label, "0", "no autocorrelation")

# Extract experiment run IDs
datin$Exp_run <- paste0(datin$Mean_temperature, str_sub(datin$Profile_name, -2, -1))
datin$Exp_run <- gsub('_', '', datin$Exp_run)
datin$Exp_run <- as.numeric(datin$Exp_run)

# Calculate total frond count
datin$Frond_count <- datin$Frond_count_1 + datin$Frond_count_2 + datin$Frond_count_3

# Remove data for which we did not calculate surface area
datin <- datin[!is.na(datin$Initial_area),]
datin <- datin[!is.na(datin$Final_area),]
datin <- subset(datin, !(Frond_count > 0 & Final_area == 0))
```


```{r include=FALSE}
## Subset Data
# We subset the data to retain only those experiment runs that have at least 3  replicates within each `Mean_Temperature` and `Species` level

datin <- datin[
   ave(seq_along(datin$Exp_run), 
       interaction(datin$Exp_run, datin$Mean_temperature, datin$Species), 
       FUN = length) >= 3, 
]
```

```{r include=FALSE}
## Further Data Processing

# Blending 10C and 15C, as there are no significant differences between performance in these 2 average temperatures
datin$Mean_temperature <- str_replace(datin$Mean_temperature, "15", "10")
datin$Mean_temperature <- str_replace(datin$Mean_temperature, "10", "10-15")

# Removing "lab" strain, as we did not separate generations for them and this had a significant effect at this mean temperature
datin <- subset(datin, !(Species == "Lab_LM" & Mean_temperature == "27"))

# Blending common duckweed strains, as preliminary analyses showed no differences between them
datin$Species <- gsub("Lab_LM|Field_LM", "LM", datin$Species)
```


## Surface area measurement used: change in surface area

```{r echo=FALSE}
dataset <- datin
dataset$Exp_run <- as.factor(dataset$Exp_run)

dataset$Mean_temperature <- as.factor(dataset$Mean_temperature)
dataset$Treatment <- as.factor(dataset$Treatment)

dataset <- within(dataset, Mean_temperature <- relevel(factor(Mean_temperature), ref = "10-15"))
dataset <- within(dataset, Treatment <- relevel(factor(Treatment), ref = "no autocorrelation"))
```

## Histogram of data (SA = change in surface area)
```{r echo=FALSE}
dataset$SA <- dataset$Final_area - dataset$Initial_area
# For each species
invisible(sapply(c("LM", "LP"), function(sp) {
  
  # Create a 3x3 grid for plotting (3 treatments x 3 mean temperatures)
  par(mfrow = c(3, 3))
  
  # Set the desired order of mean temperatures
  mean_temp_order <- c("10-15", "27", "37")
  
  # Loop through each combination of treatment and mean temperature (in specified order)
  invisible(sapply(unique(dataset$Treatment), function(trt) {
    invisible(sapply(mean_temp_order, function(temp) {
      
      sa_values <- dataset$SA[dataset$Species == sp & dataset$Treatment == trt & dataset$Mean_temperature == temp]
      min_sa <- min(sa_values, na.rm = TRUE)
      max_sa <- max(sa_values, na.rm = TRUE)
      
      # Plot histogram for each combination of species, treatment, and mean temperature
      hist(dataset$SA[dataset$Species == sp & dataset$Treatment == trt & dataset$Mean_temperature == temp],
           main = paste("Species:", sp, "\nTreatment:", trt, "\nTemperature:", temp),  
           xlab = "Surface area", col = "lightblue", border = "white",
           breaks = seq(min_sa, max_sa + 0.3, by = 0.3))
    }))
  }))
  
}))
```

## View dataset and response variable (SA = change in surface area)
```{r echo=FALSE}
knitr::kable(head(format(dataset[, c(34, 33, 18, 19, 35, 36)], digits = 2, nsmall = 3)))
```

## View number of replicates
LM
```{r echo=FALSE}
table(dataset$Mean_temperature[dataset$Species == "LM"], dataset[, 33][dataset$Species == "LM"])
```

LP
```{r echo=FALSE}
table(dataset$Mean_temperature[dataset$Species == "LP"], dataset[, 33][dataset$Species == "LP"])
```

## Mixed-Effects Model Fitting
```{r}
species_list <- c("LM", "LP")
model_results <- list()

for (sp in species_list) {
  species_data <- subset(dataset, Species == sp)
  library(lme4)
simple <- lm(SA ~ Treatment * Mean_temperature, 
                 data = species_data, 
                 na.action = na.exclude)

exp_number <- lmer(SA ~ Treatment * Mean_temperature + (1 | Exp_run), 
                         data = species_data, 
                         na.action = na.exclude)

#anova(exp_number, simple, test = "Chisq")
L0 <- logLik(simple)
L1 <- logLik(exp_number)
L.ratio <- as.vector(- 2 * (L0 - L1))
p_value <- 0.5 * (1 - pchisq(L.ratio, 1))
options(scipen = 999)
cat("Comparing simple x mixed model for ", sp, "\n")
cat("p_value:", p_value, "\n")

model_results[[sp]] <- list(simple = simple, exp_number = exp_number)

}
```

## Model Comparison and Results

We compare the performance of the models and get results for best model.

```{r echo=FALSE}
compare_results <- list()
best_models <- list()

for (sp in species_list) {
  Cand.modsF <- list("no random effects" = model_results[[sp]]$simple,
                     "experiment number" = model_results[[sp]]$exp_number)
  
  # Get AIC values for each model using base R's AIC function
  AIC_values <- sapply(Cand.modsF, AIC)
  
  # Store the AIC values in a dataframe for easy viewing
  compareF <- data.frame(
    Name = names(AIC_values),
    AIC = AIC_values
  )
  
  # Display AIC values in the dataframe (for each species)
  compare_results[[sp]] <- knitr::kable(compareF, "simple", digits=2)
  
  # Get the model name with the lowest AIC
  best_model_name <- compareF$Name[which.min(compareF$AIC)]
  best_models[[sp]] <- Cand.modsF[[best_model_name]]
  
  # Display model comparison for the current species
  print(paste("Model comparison for species:", sp))
  print(compare_results[[sp]])
  cat("\n")
   # Display model summary for the current species
  print(paste("Model summary for species:", sp))
  best_model <- best_models[[sp]]
  print(best_model)
  cat("\n")
}
```

## Model significance testing
The best model selected in the step above is tested for significance, for each species
```{r echo=FALSE}
library(car)

for (sp in species_list) {
  best_model <- best_models[[sp]]
  
    # Type III Wald Chi-Square Tests (for interactions)
    anova_output <- car::Anova(best_model, type = "III")
    

    # Add p-values 
    if ("F value" %in% colnames(anova_output)) {
      anova_output$`Pr(>F)` <- pf(anova_output$`F value`, 
                                   anova_output$Df, 
                                   df.residual(best_model), 
                                   lower.tail = FALSE)
    }

  print(paste("Predictor significance - species", sp, ":"))
  print(anova_output)
  
  residual_df <- df.residual(best_model)
  print(paste("Residual degrees of freedom - species :", sp, ":", residual_df))
}

  print(paste("Showing predictor significance for linear mixed model for LM"))
    lpmixed <- model_results[["LM"]]$exp_number
    print(car::Anova(lpmixed, type = "III"))
    
  print(paste("Showing predictor significance for linear mixed model for LP"))
    lpmixed <- model_results[["LP"]]$exp_number
    print(car::Anova(lpmixed, type = "III"))
    
```

## Showing results for linear mixed model for common duckweeds (treatment is significant)
## R2 for common duckweeds
```{r echo=FALSE}
library(sjPlot)
best_model <- model_results[["LM"]]$exp_number
summary_results <- tab_model(best_model)
print(paste("Summary for FLM:"))
summary_results
```

## Model diagnostics

```{r model-validation, echo=FALSE, fig.show='hold'}
  best_model <- model_results[["LM"]]$exp_number
  
 # Plot residual diagnostics
  print(plot(best_model, main = paste("Residuals vs Fitted")))
  
  # Q-Q plot for normality of residuals
  qqnorm(residuals(best_model), main = paste("Q-Q Plot"))
  qqline(residuals(best_model), col = "red")
  
# Perform a Shapiro-Wilk test on random effects for normality
  # ranef_component <- ranef(best_model)
  # print(shapiro.test(ranef_component[[1]][, "(Intercept)"]))

```

## Post-hoc test for common duckweeds
```{r echo=FALSE}
  library(emmeans)
  best_model <- model_results[["LM"]]$exp_number
   posthoc <- emmeans(best_model, pairwise ~ Treatment | Mean_temperature,  lmer.df="satterthwaite") 
   print(paste("Pairwise interactions:"))
   print(posthoc)
```

```{r echo=FALSE}
library(ggplot2)
dataset_species <- subset(dataset, Species == "LM")
summary_data <- aggregate(SA ~ Treatment + Mean_temperature, data = dataset_species, FUN = function(x) c(mean = mean(x, na.rm = TRUE), se = sd(x, na.rm = TRUE) / sqrt(length(x))))
summary_data <- do.call(data.frame, summary_data)

plot <- ggplot(summary_data, aes(x = Mean_temperature, y = SA.mean, 
                                 color = Treatment, group = Treatment)) +
  geom_errorbar(aes(ymin = SA.mean - SA.se, ymax = SA.mean + SA.se), 
                width = 0.3, linewidth = 1.5, position = position_dodge(width = 0.8)) +  
  geom_point(size = 4, position = position_dodge(width = 0.8)) +  
  scale_color_manual(values = c("no autocorrelation" = "grey37", 
                                "cold-hot" = "#0072B2", 
                                "hot-cold" = "#D55E00")) +
  theme_bw(base_size = 24) +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        legend.position = "none",
        axis.line = element_line(colour = "black")) +
  labs(y = expression(Delta~"surface area (cm"^2*")"), x = expression("Mean temperature (" * degree * "C)")) +  
  guides(color = guide_legend(title = "Treatment"), linetype = guide_legend(title = "Treatment")) +
  ylim(-0.1, 3) +
  
  geom_text(aes(x = 3.25, y = 0.7, label = "*"), color = "black", vjust = -0.5, size = 7) +
  
  geom_segment(aes(x = 3.05, y = 0.8, xend = 2.65, yend = 0.8), color = "black", size = 1) +
  geom_segment(aes(x = 3.3, y = 0.8, xend = 3.2, yend = 0.8), color = "black", size = 1)

plot
```


