---
title: "Survival analysis"
author: "Debora"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


```{r, include=FALSE, warning=F, message=F}
library(tidyverse)
library(broom)
library(cowplot)
library(ggthemes)

datin1 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/metafile_autocorrelation_2022-2023.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")

datin2 <- read.csv("https://raw.githubusercontent.com/Cuddington-Lab/thermal-experiments/main/metafile_autocorrelation_2021-2022.csv",header=TRUE, stringsAsFactors = TRUE,fileEncoding="UTF-8-BOM")

datin <- rbind(datin1,datin2)

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9))
              &!(datin$Treatment == 0.95 & (datin$Obs_sd <= 2.1 | datin$Obs_sd >= 2.9)),]

datin <- datin[!(datin$Treatment == 0 & (datin$Obs_ac <= -0.2 | datin$Obs_ac >= 0.2))
               &!(datin$Treatment == 0.95 & (datin$Obs_ac <= 0.92 | datin$Obs_ac >= 0.98)),]

datin <- datin[!datin$Experiment_Number == 83,]
datin <- datin[!(datin$Experiment_Number == 99 & datin$Incubator ==3),]
datin <- datin[!(datin$Experiment_Number == 99 & datin$Incubator ==4),]

datin <- subset(datin, Mean_temperature == 37)

#Create new treatment label (cat_1_4: if investigating effect of initial 
#sequence slope; cat_1: whole sequence)
table(datin$Treatment, datin$cat_1)
levels(datin$cat_1) = c("N/A","N","P","N/A")
datin$label<-paste0(datin$Treatment, datin$cat_1)
table(datin$label)

datin$Frond_count <- datin$Frond_count_1 + datin$Frond_count_2 + datin$Frond_count_3
datin$Mortality <- datin$Dead_frond_count_1 + datin$Dead_frond_count_2 + datin$Dead_frond_count_3
```

### Identifying surviving and non-surviving plants

Plants were considered as dead when they displayed >90% loss of green pigmentation at the end of the experiment (Thomson & Dennis, 2013).

The original result dataset for the autocorrelation experiment was modified so that each row (or each replicate) was repeated to correspond to the amount of living and dead individuals. Each non-surviving individual frond received a code of "0" (zero), and each surviving individual frond received a code of "1" (one). Each line below corresponds to a single individual frond.
```{r, include=FALSE}
dish <- as.data.frame(lapply(datin, rep, datin$Frond_count_1))
dish$dish <- rep("d1",times=468)
dish$Frond_count <- dish$Frond_count_1

dish2 <- as.data.frame(lapply(datin, rep, datin$Frond_count_2))
dish2$dish <- rep("d2",times=431)
dish2$Frond_count <- dish2$Frond_count_2

dish3 <- as.data.frame(lapply(datin, rep, datin$Frond_count_3))
dish3$dish <- rep("d3",times=430)
dish3$Frond_count <- dish3$Frond_count_3
living <- rbind(dish,dish2,dish3)

dish <- as.data.frame(lapply(datin, rep, datin$Dead_frond_count_1))
dish$dish <- rep("d1",times=97)
dish$Mortality <- dish$Dead_frond_count_1

dish2 <- as.data.frame(lapply(datin, rep, datin$Dead_frond_count_2))
dish2$dish <- rep("d2",times=90)
dish2$Mortality <- dish2$Dead_frond_count_2

dish3 <- as.data.frame(lapply(datin, rep, datin$Dead_frond_count_3))
dish3$dish <- rep("d3",times=77)
dish3$Mortality <- dish3$Dead_frond_count_3
mort <- rbind(dish,dish2,dish3)

#mort <- as.data.frame(lapply(datin, rep, datin$Mortality))
mort <- subset(mort, !mort$Mortality == 0)
mort["Mortality"][mort["Mortality"] >= 1] <- 0
names(mort)[names(mort) == 'Mortality'] <- 'survival'
mort <- subset(mort, select = -(Frond_count))

#living <- as.data.frame(lapply(datin, rep, datin$Frond_count))
living <- subset(living, !living$Frond_count == 0)
living["Frond_count"][living["Frond_count"] >= 1] <- 1
names(living)[names(living) == 'Frond_count'] <- 'survival'
living <- subset(living, select = -(Mortality))
datin3 <- rbind(living,mort)

datin3$Species <- as.factor(datin3$Species)
datin3$label <- as.factor(datin3$label)
datin3$dish <- as.factor(datin3$dish)
```

```{r}
head(datin3[c(1,5,6,7,8,33)],3)
tail(datin3[c(1,5,6,7,8,33)],3)
```


```{r, echo=FALSE, warning=FALSE, messages=FALSE}
library(stringr)
datin3$label <- str_replace(datin3$label, "0N/A", "0")
datin3$label <- str_replace(datin3$label, "0.95N", "hot-cold")
datin3$label <- str_replace(datin3$label, "0.95P", "cold-hot")

datin4 <- subset(datin3, Species == "Field_LM")

library(ggplot2)
lm <- ggplot(datin4, aes(x = label, y = survival, color = label)) + 
  geom_jitter() +
  labs(title = "L. minor") +
  xlab("treatment") +
  geom_smooth(formula = y ~ x,method=lm,se=T)+ theme_bw() + theme(panel.border = element_blank(), 
                                                  panel.grid.major = element_blank(),
                                                  panel.grid.minor = element_blank(), 
                                                  axis.line = element_line(colour = "black"),
                                                  legend.position="none",
                                                  axis.text=element_text(size=15),
                                                  axis.title=element_text(size=15),
                                                  legend.text=element_text(size=13),
                                                  plot.title = element_text(face = "italic")) 

datin5 <- subset(datin3, Species == "LP")
lp <- ggplot(datin5, aes(x = label, y = survival, color = label)) + 
  geom_jitter() +
  labs(title = "L. punctata") +
  xlab("treatment") +
  geom_smooth(formula = y ~ x,method=lm,se=T)+ theme_bw() + theme(panel.border = element_blank(), 
                                                  panel.grid.major = element_blank(),
                                                  panel.grid.minor = element_blank(), 
                                                  axis.line = element_line(colour = "black"),
                                                  legend.position="none",
                                                  axis.text=element_text(size=15),
                                                  axis.title=element_text(size=15),
                                                  legend.text=element_text(size=13),
                                                  plot.title = element_text(face = "italic"))

library("gridExtra")
grid.arrange(lm, lp, ncol = 2) 
```

### Number of observations
```{r, echo=F, warning=F, comment = NA, message=F}
knitr::kable(table(datin3$Species, datin3$survival),"simple")

table(datin$Species,datin$label)
```

### Finding the best model for the survival data:
```{r message=FALSE, warning=FALSE, comment=NA, include=FALSE}
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(AICcmodavg) #https://www.montana.edu/rotella/documents/502/FawnsinR.html

#Logistic regression allows categorically and continuously scaled variables to predict any categorically scaled criterion
#Osborne, J. W. (Ed.). (2008). Best practices in quantitative methods. Sage.

# adding random effects
library(lme4)
glm.fit.sp = glm(survival ~ label*Species, family = binomial, data=datin3)
experim <- glmer(survival ~ label*Species + (1|Experiment_Number), family=binomial, data=datin3)
incub <- glmer(survival ~ label*Species + (1|Incubator), family=binomial, data=datin3)
dish <- glmer(survival ~ label*Species + (1|dish), family=binomial, data=datin3)
experim_inc_dish <- glmer(survival ~ label*Species + (1|Experiment_Number) + (1|Incubator) + (1|dish), family=binomial, data=datin3)
inc_dish <- glmer(survival ~ label*Species + (1|Incubator)+(1|dish), family=binomial, data=datin3)
inc_exp <- glmer(survival ~ label*Species + (1|Incubator)+(1|Experiment_Number), family=binomial, data=datin3)
dish_exp <- glmer(survival ~ label*Species + (1|dish)+(1|Experiment_Number), family=binomial, data=datin3)
exp.inc_dish <- glmer(survival ~ label*Species + (1|Experiment_Number/Incubator) + (1|dish), family=binomial, data=datin3)
exp.inc <- glmer(survival ~ label*Species + (1|Experiment_Number/Incubator), family=binomial, data=datin3)

#odds ratio: probability of success / probability of failure
#for categorical features or predictors, the odds ratio compares the odds of the event occurring for each category of the predictor relative to the reference category, given that all other variables remain constant

#model assessment
##set up named list
Cand.mods <- list("no random effects" = glm.fit.sp,
                  "experiment number" = experim,
                  "incubator" = incub, 
                  "dish" = dish,
                  "incubator, dish, and experiment number" = experim_inc_dish,
                  "incubator and dish" = inc_dish,
                  "incubator and experiment number" = inc_exp,
                  "dish and experiment number" = dish_exp,
                  "incubator nested in experiment number, and dish" = exp.inc_dish,
                  "incubator nested in experiment number" = exp.inc)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NA}
#compute table
library(performance)
compare <- compare_performance(Cand.mods)
knitr::kable(compare[,c(1,2,3,4)],"simple")


# Akaike information criteria (AIC): allows for comparing different models 
# AIC considers both model fit and parsimony (lowest # of parameters)
# we are looking for lowest value of the AIC metric, since the complexity penalty makes AIC larger
# AICs within 2 units: no importantly different predictive performance

# AIC_wt: Akaike weights; probability that a given model is the best one, 
# given the data and the set of candidate models (Burnham & Anderson, 2001)
# differences in AIC with respect to the AIC of the best 
# candidate model; relative model likelihoods divided by sum of 
# likelihoods of all models to obtain Akaike weights
# 0.73/0.26=2.80: best model is 2.8x times than 2nd best model
```

### Fitting a logistic model to the survival data:
```{r echo=F, message=FALSE, warning=FALSE, comment=NA}
#I can use marginal/conditional r2 for glmm binomial: http://rstudio-pubs-static.s3.amazonaws.com/13853_668273e779754a579aa535d4d11d72c8.html
tab_model(exp.inc)
```

The hot-cold strong autocorrelation treatment is significantly different than the control (zero autocorrelation). The interaction effect between the cold-hot strong autocorrelation treatment and the invasive species (LP) is also significant.

```{r, include=F}
glm.probs <- predict(exp.inc,datin3, type = "response")
```

### Plotting estimates of the glm:  
* the thick line (vertical intercept at position 1) indicates no effect  
* horizontal lines are error bars  
* numbers are estimates values  
* asterisks are significance levels of p-values  
```{r message=FALSE, warning=FALSE}
library(sjPlot)
plot_model(exp.inc,vline.color = "grey",show.values = TRUE, value.offset = .3) + 
                                                  theme_bw() + 
                                                  theme(panel.border = element_blank(), 
                                                  panel.grid.major = element_blank(),
                                                  panel.grid.minor = element_blank(), 
                                                  axis.line = element_line(colour = "black"),
                                                  axis.text=element_text(size=15),
                                                  axis.title=element_text(size=15),
                                                  legend.text=element_text(size=13))

set_theme(base = theme_light())
estimates2 <- plot_model(exp.inc, show.values = TRUE, value.offset = 0.3,
           title = "Change in survival as compared to control",vline.color = "red") + 
  theme(text = element_text(size = 20))

# vline
estimates2$layers[[1]]$aes_params$size <- 4
# labels
estimates2$layers[[4]]$aes_params$size <- 8

estimates2
```

### Plotting model predictions (considering interactions terms):  
* Vertical lines are error bars  
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(sjPlot)
plot_model(exp.inc, type = "int", se=T, line.size = 2) + theme_bw() + theme(panel.border = element_blank(), 
                                                  panel.grid.major = element_blank(),
                                                  panel.grid.minor = element_blank(), 
                                                  axis.line = element_line(colour = "black"),
                                                  #legend.position="none",
                                                  axis.text=element_text(size=16),
                                                  axis.title=element_text(size=16),
                                                  plot.title = element_text(size=20),
                                                  legend.text=element_text(size=14),
                                                  legend.title=element_text(size=15)) +
                                                  labs(y= "Survival", x = "Treatment") +
                                                  scale_color_manual(values = c("Field_LM" = "#999999", 
                                                                                "LP" = "#E69F00"))
#info on error bars (standard error of the means): 
# https://www.graphpad.com/support/faq/spanwhat-you-can-conclude-when-two-error-bars-overlap-or-dontspan/
```

 
### Confusion matrix: determining model occurrence predictions based on threshold (eg., 60% survival); left diagonal has correctly predicted values
```{r, echo=FALSE, warning=FALSE, message=FALSE}
logocc <- ifelse(glm.probs > 0.6, 1, 0)

# Calculate a confusion matrix
ctab = table(logocc, datin3$survival)
dimnames(ctab) <- list(Actual = c("absence(0)", "presence(1)"),
                       Predicted = c("absence(0)", "presence(1)"))
ctab
```

```{r, include=FALSE}
#for 0.5 threshold
#if we had 91 true negatives (TN), 63 false positives (FP), 139 false negatives (FN), and 1020 true positives (TP):
tn=91
fp=100
fn=139
tp=1107
error_rate = (fn+fp)/(tn+fn+tp+fp)
#error_rate = 15%
(sum(tn,fp,fn,tp))
sensitivity = 1-(fp/(sum(tn,fp,fn,tp)))
specificity = tn/(tn+fp)
```

### Making a receiver operator curve (ROC) to analyze the model performance across several thresholds 
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ROCit)
ROCit_obj <- rocit(score = glm.fit.sp$fitted.values, class = datin3$survival)
plot(ROCit_obj)
```

### Obtaining area under the curve (maximum is one) 
```{r,echo=FALSE, warning=FALSE, message=FALSE}
summary(ROCit_obj)
```

### Optimal threshold value given by the Youden index 
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(pROC)
roc_obj <- roc(datin3$survival, glm.probs)
coords(roc_obj, "best", "threshold")
```

### Post-hoc comparisons
```{r,echo=FALSE,warning=FALSE, message=FALSE}
library(emmeans) #emmeans has been used for GLMER & binary data (internet forums)
pairwise <- emmeans(exp.inc, list(pairwise ~ label*Species), lmer.df="satterthwaite")
pairwise
pairs <- as.data.frame(pairwise$`pairwise differences of label, Species`)
pairs <- subset(pairs,p.value<=0.05)
pairs
```


References:  

* https://www.quantitative-biology.ca/

* (Thomson & Dennis, 2013): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0080102

* https://strengejacke.github.io/sjPlot/articles/plot_model_estimates.html#fitting-a-logistic-regression-model
